---
title: "Measles Watch"
author: "100318577"
format: 
  dashboard:
    theme: pulse
    scrolling: true
---

```{=html}
<style>
.navbar {
  min-height: 60px; /* Adjust the height */
}
.navbar a {
  font-size: 20px; /* Increase font size */
  font-weight: bold; /* Make text bold */
  color: white !important; /* Set text color to white */
}
.tab-content {
  font-weight: bold !important;
  font-size: 20px !important;
}
.valuebox {
  max-height: 200px !important;
  height: 190px !important;
}
.valuebox .value {
  font-size: 25px !important;
}
.valuebox .card-body {
  font-size: 27px !important;
}
.valuebox h1, .valuebox h2, .valuebox h3 {
  font-size: 35px !important;
}
.valuebox * {
  font-size: 25px !important;
}
/* Increase all value box icons to fill 75% of left space - multiple selectors to ensure it works */
.bslib-value-box .bslib-value-box-icon,
.valuebox .icon,
.value-box-icon,
[class*="value-box"] .icon,
[class*="valuebox"] .icon {
  font-size: 75px !important; /* Increased to fill 75% of left space */
}
/* Additional fallback for Bootstrap icons */
.bi {
  font-size: 64px !important;
}
/* Set all value box text and icons to white */
.bslib-value-box .bslib-value-box-icon {
  color: white !important;
}
.bslib-value-box .bslib-value-box-title {
  color: white !important;
}
.bslib-value-box .bslib-value-box-value {
  color: white !important;
}
.bslib-value-box * {
  color: white !important;
}
</style>
```

```{r}
#| include: false
#| lable: load_global_data
#| message: false

library(tidyverse)
library(plotly)
library(readr)
library(readxl)
library(DT)
library(maps)
library(sf)
library(htmlwidgets)
library(leaflet)

## GLOBAL DATA ----
### Country coordinates from the maps package ----
country_coords <- world.cities %>%
  group_by(country.etc) %>%
  summarise(
    lat = mean(lat, na.rm = TRUE),
    lng = mean(long, na.rm = TRUE)
  ) %>%
  rename(country = country.etc)

### Global cases and deaths ----
# 🚀 Load & Process Global Measles Data (1980–2021)
global_cases_deaths_summary <- read_csv("data/measles_global/global_cases_deaths.csv") %>%
  select(measure, location, year, val) %>%  # ✅ Keep only relevant columns
  filter(measure %in% c("Incidence", "Deaths")) %>%  # ✅ Focus only on cases & deaths
  group_by(location, year, measure) %>%
  summarise(Value = round(sum(val, na.rm = TRUE), 0), .groups = "drop") %>%  # ✅ Round values for clean display
  pivot_wider(names_from = measure, values_from = Value, values_fill = 0) %>%  # ✅ Separate cases & deaths
  rename(cases = Incidence, deaths = Deaths) %>%  # ✅ Rename for clarity
  left_join(country_coords, by = c("location" = "country")) %>%  # ✅ Merge with country coordinates
  filter(year >= 1980 & year <= 2021) %>%  # ✅ Ensure only years within 1980-2021 are included
  group_by(location, year) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    total_deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop")

### Global cases by sex and age ----
global_sex_age_grouped <- read_csv("data/measles_global/global_cases_sex_age.csv") %>%
  select(location, sex, age, year, val) %>%  # ✅ Removed 'measure' column
  filter(sex %in% c("Male", "Female")) %>%
  mutate(
    age_group = case_when(
      age == "<1 year" ~ "<1",
      age == "2-4 years" ~ "2-4",
      age %in% c("5-14 years", "15-19 years") ~ "5-19",
      age %in% c("20-24 years", "25-29 years") ~ "20-29",
      age %in% c("30-34 years", "35-39 years") ~ "30-39",
      age %in% c("40-44 years", "45-49 years") ~ "40-49",
      age %in% c("50-54 years", "55-59 years") ~ "50-59",
      age %in% c("60-64 years", "65-69 years") ~ "60-69",
      age == "70+ years" ~ "70+",
      TRUE ~ "no_data"  # ✅ Explicitly marking missing age group as "no_data"
    )
  ) %>%
  group_by(location, year, sex, age_group) %>%
  summarise(cases = round(sum(val, na.rm = TRUE), 0), .groups = "drop") %>%
  left_join(country_coords, by = c("location" = "country")) %>%
  filter(year >= 1980 & year <= 2021) %>%
  mutate(
    lat = replace_na(lat, 0),
    lng = replace_na(lng, 0)
  )

# CREATE CONTINENT MAPPING AND VARIABLES NEEDED FOR VALUE BOXES ----
# Define color palette for continents
colors_vector <- c(
  "Africa" = "#65463E",   # Carafe, representing the rich wildlife and savannah
  "Asia" = "#0A7029",     # Green, symbolizing lush forests and growth
  "Europe" = "#2535D9",   # Royal Blue, echoing tradition and stability
  "Americas" = "#B22222",  # Firebrick Red, conveying energy and history
  "Oceania" = "#C08D2C"   # Desert Sun, reflecting arid landscapes and warmth
)

# Updated continent mapping
continent_mapping <- data.frame(
  location = c("Afghanistan", "Albania", "Algeria", "Angola", "Argentina", "Armenia", "Australia", 
               "Austria", "Azerbaijan", "Bahrain", "Bangladesh", "Belarus", "Belgium", "Benin", 
               "Bolivia", "Bosnia and Herzegovina", "Botswana", "Brazil", "Bulgaria", "Burkina Faso",
               "Burundi", "Cambodia", "Cameroon", "Canada", "Central African Republic", "Chad", 
               "Chile", "China", "Colombia", "Comoros", "Congo", "Costa Rica", "Croatia", "Cuba",
               "Cyprus", "Czech Republic", "Democratic Republic of Congo", "Denmark", "Djibouti",
               "Dominican Republic", "Ecuador", "Egypt", "El Salvador", "Equatorial Guinea", 
               "Eritrea", "Estonia", "Eswatini", "Ethiopia", "Finland", "France", "Gabon", 
               "Gambia", "Georgia", "Germany", "Ghana", "Greece", "Guatemala", "Guinea", 
               "Guinea-Bissau", "Guyana", "Haiti", "Honduras", "Hungary", "Iceland", "India", 
               "Indonesia", "Iran", "Iraq", "Ireland", "Israel", "Italy", "Jamaica", "Japan", 
               "Jordan", "Kazakhstan", "Kenya", "Kuwait", "Kyrgyzstan", "Laos", "Latvia", 
               "Lebanon", "Lesotho", "Liberia", "Libya", "Lithuania", "Luxembourg", "Madagascar", 
               "Malawi", "Malaysia", "Maldives", "Mali", "Malta", "Mauritania", "Mauritius", 
               "Mexico", "Moldova", "Mongolia", "Montenegro", "Morocco", "Mozambique", "Myanmar", 
               "Namibia", "Nepal", "Netherlands", "New Zealand", "Nicaragua", "Niger", "Nigeria", 
               "North Korea", "North Macedonia", "Norway", "Oman", "Pakistan", "Panama", 
               "Papua New Guinea", "Paraguay", "Peru", "Philippines", "Poland", "Portugal", 
               "Qatar", "Romania", "Russia", "Rwanda", "Saudi Arabia", "Senegal", "Serbia", 
               "Sierra Leone", "Singapore", "Slovakia", "Slovenia", "Somalia", "South Africa", 
               "South Korea", "South Sudan", "Spain", "Sri Lanka", "Sudan", "Suriname", "Sweden", 
               "Switzerland", "Syria", "Tajikistan", "Tanzania", "Thailand", "Timor-Leste", 
               "Togo", "Tunisia", "Turkey", "Turkmenistan", "Uganda", "Ukraine", "United Arab Emirates", 
               "United Kingdom", "United States", "Uruguay", "Uzbekistan", "Venezuela", "Vietnam", 
               "Yemen", "Zambia", "Zimbabwe"),
  
  continent = c("Asia", "Europe", "Africa", "Africa", "Americas", "Asia", "Oceania",
                "Europe", "Asia", "Asia", "Asia", "Europe", "Europe", "Africa",
                "Americas", "Europe", "Africa", "Americas", "Europe", "Africa",
                "Africa", "Asia", "Africa", "Americas", "Africa", "Africa",
                "Americas", "Asia", "Americas", "Africa", "Africa", "Americas", "Europe", "Americas",
                "Europe", "Europe", "Africa", "Europe", "Africa",
                "Americas", "Americas", "Africa", "Americas", "Africa",
                "Africa", "Europe", "Africa", "Africa", "Europe", "Europe", "Africa",
                "Africa", "Asia", "Europe", "Africa", "Europe", "Americas", "Africa",
                "Africa", "Americas", "Africa", "Americas", "Europe", "Europe", "Asia",
                "Asia", "Asia", "Asia", "Europe", "Asia", "Europe", "Americas", "Asia",
                "Asia", "Asia", "Africa", "Asia", "Asia", "Asia", "Europe",
                "Asia", "Africa", "Africa", "Africa", "Europe", "Europe", "Africa",
                "Africa", "Asia", "Asia", "Africa", "Europe", "Africa", "Africa",
                "Americas", "Europe", "Asia", "Europe", "Africa", "Africa", "Asia",
                "Africa", "Asia", "Europe", "Oceania", "Americas", "Africa", "Africa",
                "Asia", "Europe", "Europe", "Asia", "Asia", "Americas",
                "Oceania", "Americas", "Americas", "Asia", "Europe", "Europe",
                "Asia", "Europe", "Europe", "Africa", "Europe", "Asia", "Africa",
                "Africa", "Asia", "Europe", "Europe", "Africa", "Africa",
                "Asia", "Africa", "Europe", "Asia", "Africa", "Americas", "Europe",
                "Europe", "Asia", "Asia", "Africa", "Asia", "Asia",
                "Africa", "Africa", "Asia", "Asia", "Africa", "Europe", "Asia",
                "Europe", "Americas", "Americas", "Asia", "Asia", "Asia",
                "Europe", "Africa", "Africa")
)

# Prepare country-level data for cases
country_cases_data <- global_cases_deaths_summary %>%
  filter(!location %in% c("Africa", "Asia", "Europe", "Americas", "Oceania", "World")) %>%
  filter(total_cases > 0) %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent))

# Get available years and continents
years_available <- sort(unique(country_cases_data$year))
continents_available <- sort(unique(country_cases_data$continent))

# CREATE CONTINENTAL DATA FOR CHARTS
continent_cases <- global_cases_deaths_summary %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent)) %>%
  group_by(continent, year) %>%
  summarise(
    total_cases = sum(total_cases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    cases_millions = total_cases / 1000000,
    exact_cases = total_cases
  )

continent_deaths <- global_cases_deaths_summary %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent)) %>%
  group_by(continent, year) %>%
  summarise(
    total_deaths = sum(total_deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    deaths_thousands = total_deaths / 1000,
    exact_deaths = total_deaths
  )

# Define continent colors
continent_colors <- colors_vector

# Read vaccination data
vac_1dose <- read_csv("data/measles_global/1dose_share-of-children-vaccinated-against-measles.csv") %>%
  rename(`1_dose` = `Share of one-year-olds who have received their first dose of measles-containing vaccine (MCV1)`,
         `location` = `Entity`, `year` = `Year`) %>%
  left_join(country_coords, by = c("location" = "country"))

vac_2dose <- read_csv("data/measles_global/2_doses_ share-of-children-vaccinated-with-mcv.csv") %>%
  rename(`2_dose` = `Share of children who have received two doses of measles-containing vaccine as per the national schedule (MCV2)`,
         `location` = `Entity`, `year` = `Year`) %>%
  left_join(country_coords, by = c("location" = "country"))

start_year_1dose <- min(vac_1dose$year, na.rm = TRUE)
start_year_2dose <- min(vac_2dose$year, na.rm = TRUE)
later_start_year <- max(start_year_1dose, start_year_2dose)

vac_1dose_filtered <- vac_1dose %>% filter(year >= later_start_year)
vac_2dose_filtered <- vac_2dose %>% filter(year >= later_start_year)

vac_combined <- vac_1dose_filtered %>%
  left_join(vac_2dose_filtered %>% select(location, year, `2_dose`), by = c("location", "year")) %>%
  rename(dose_1 = `1_dose`, dose_2 = `2_dose`)
```

# Global

## GLOBAL ROW 1 - Global boxes {height="200px"}

```{r}
#| include: false
#| lable: global_boxes
#| message: false

# Filter data based on selected continents
filtered_data <- country_cases_data %>%
  filter(continent %in% continents_available)

# Create a data frame for all years
stats_data <- map_dfr(years_available, function(yr) {
  year_data <- filtered_data %>% filter(year == yr)
  # Total cases for the year
  total_cases_year <- sum(year_data$total_cases, na.rm = TRUE)
  # Country with highest cases
  highest_country <- year_data %>%
    arrange(desc(total_cases)) %>%
    slice_head(n = 1)
  # Total deaths for the year
  total_deaths_year <- sum(year_data$total_deaths, na.rm = TRUE)
  # Country with highest deaths
  highest_deaths_country <- year_data %>%
    arrange(desc(total_deaths)) %>%
    slice_head(n = 1)
  data.frame(
    year = yr,
    total_cases = total_cases_year,
    highest_cases_country = if(nrow(highest_country) > 0) highest_country$location[1] else NA,
    highest_cases_continent = if(nrow(highest_country) > 0) highest_country$continent[1] else NA,
    highest_cases_value = if(nrow(highest_country) > 0) highest_country$total_cases[1] else 0,
    total_deaths = total_deaths_year,
    highest_deaths_country = if(nrow(highest_deaths_country) > 0) highest_deaths_country$location[1] else NA,
    highest_deaths_continent = if(nrow(highest_deaths_country) > 0) highest_deaths_country$continent[1] else NA,
    highest_deaths_value = if(nrow(highest_deaths_country) > 0) highest_deaths_country$total_deaths[1] else 0
  )
})

# Find the earliest and latest year
min_year <- min(stats_data$year, na.rm = TRUE)
max_year <- max(stats_data$year, na.rm = TRUE)

# Extract stats for both years
min_year_stats <- stats_data %>% filter(year == min_year) %>%
  select(highest_cases_country, highest_deaths_country, total_cases, total_deaths)
max_year_stats <- stats_data %>% filter(year == max_year) %>%
  select(highest_cases_country, highest_deaths_country, total_cases, total_deaths)

# Calculate percentage change
cases_pct_change <- ((max_year_stats$total_cases - min_year_stats$total_cases) / min_year_stats$total_cases) * 100
deaths_pct_change <- ((max_year_stats$total_deaths - min_year_stats$total_deaths) / min_year_stats$total_deaths) * 100

# Create a summary dataframe
percentage_change_summary <- data.frame(
  metric = c("Total Cases", "Total Deaths"),
  min_year = c(min_year_stats$total_cases, min_year_stats$total_deaths),
  max_year = c(max_year_stats$total_cases, max_year_stats$total_deaths),
  percentage_change = c(cases_pct_change, deaths_pct_change)
)

# 1) Find the country with the highest measles cases and deaths overall

# Country with highest total cases across all years
highest_cases_country_overall <- filtered_data %>%
  group_by(location, continent) %>%
  summarise(max_cases = max(total_cases, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(max_cases)) %>%
  slice_head(n = 1)

# Country with highest total deaths across all years
highest_deaths_country_overall <- filtered_data %>%
  group_by(location, continent) %>%
  summarise(max_deaths = max(total_deaths, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(max_deaths)) %>%
  slice_head(n = 1)

# 2) Find percentage difference between highest and lowest cases for the highest-case country

# Get the country name with highest cases
top_country_name <- highest_cases_country_overall$location[1]

# Get all data for this country across all years
top_country_data <- filtered_data %>%
  filter(location == top_country_name) %>%
  arrange(year)

# Find min and max values for this country
country_stats <- top_country_data %>%
  summarise(
    min_cases = min(total_cases, na.rm = TRUE),
    max_cases = max(total_cases, na.rm = TRUE),
    min_deaths = min(total_deaths, na.rm = TRUE),
    max_deaths = max(total_deaths, na.rm = TRUE),
    .groups = 'drop'
  )

# Calculate percentage differences
cases_pct_diff <- ((country_stats$max_cases - country_stats$min_cases) / country_stats$min_cases) * 100
deaths_pct_diff <- ((country_stats$max_deaths - country_stats$min_deaths) / country_stats$min_deaths) * 100

# Create summary for the top country
top_country_summary <- data.frame(
  country = top_country_name,
  continent = highest_cases_country_overall$continent[1],
  metric = c("Cases", "Deaths"),
  min_value = c(country_stats$min_cases, country_stats$min_deaths),
  max_value = c(country_stats$max_cases, country_stats$max_deaths),
  percentage_difference = c(cases_pct_diff, deaths_pct_diff)
)


# Optional: Show year-by-year data for the top country
top_country_data %>% select(year, total_cases, total_deaths)

# Additional Analysis: Top 5 countries by cases and deaths
top_5_cases <- filtered_data %>%
  group_by(location, continent) %>%
  summarise(max_cases = max(total_cases, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(max_cases)) %>%
  slice_head(n = 5)


top_5_deaths <- filtered_data %>%
  group_by(location, continent) %>%
  summarise(max_deaths = max(total_deaths, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(max_deaths)) %>%
  slice_head(n = 5)


# Analysis by continent
continent_summary <- filtered_data %>%
  group_by(continent) %>%
  summarise(
    total_cases = sum(total_cases, na.rm = TRUE),
    total_deaths = sum(total_deaths, na.rm = TRUE),
    avg_cases = mean(total_cases, na.rm = TRUE),
    avg_deaths = mean(total_deaths, na.rm = TRUE),
    countries_count = n_distinct(location),
    .groups = 'drop'
  ) %>%
  arrange(desc(total_cases))

# Summary statistics across all data
overall_stats <- filtered_data %>%
  summarise(
    total_countries = n_distinct(location),
    total_years = n_distinct(year),
    total_records = n(),
    sum_all_cases = sum(total_cases, na.rm = TRUE),
    sum_all_deaths = sum(total_deaths, na.rm = TRUE),
    avg_cases_per_record = mean(total_cases, na.rm = TRUE),
    avg_deaths_per_record = mean(total_deaths, na.rm = TRUE),
    median_cases = median(total_cases, na.rm = TRUE),
    median_deaths = median(total_deaths, na.rm = TRUE)
  )

```

### Total Measles Cases (1990-2021)

```{r}
#| content: valuebox
#| title: "Total Measles Cases"
#| icon: virus
#| color: danger
#| fig-height: 200
list(
  value = paste0(format(max_year_stats$total_cases, big.mark = ","), " (", round(cases_pct_change, 1), "% change)")
)
```

### Highest Recorded Cases

```{r}
#| content: valuebox
#| title: "Highest Recorded Cases"
#| icon: hospital
#| color: danger
#| fig-height: 200
highest_cases_year <- top_country_data %>% 
  filter(total_cases == max(total_cases, na.rm = TRUE)) %>% 
  pull(year)
highest_cases_pct <- top_country_summary %>% 
  filter(metric == "Cases") %>% 
  pull(percentage_difference)
highest_cases_value <- top_country_summary %>% 
  filter(metric == "Cases") %>% 
  pull(max_value)
list(
  value = paste0(top_country_name, " - ", format(highest_cases_value, big.mark = ","), " (", highest_cases_year, ")")
)
```

### Total Measles Deaths (1990-2021)

```{r}
#| content: valuebox
#| title: "Total Measles Deaths"
#| icon: virus
#| color: info
#| fig-height: 200
list(
  value = paste0(format(max_year_stats$total_deaths, big.mark = ","), " (", round(deaths_pct_change, 1), "% change)")
)
```

### Highest Recorded Deaths

```{r}
#| content: valuebox
#| title: "Highest Recorded Deaths"
#| icon: hospital
#| color: info
#| fig-height: 200
highest_deaths_year <- filtered_data %>% 
  filter(location == highest_deaths_country_overall$location[1]) %>% 
  filter(total_deaths == max(total_deaths, na.rm = TRUE)) %>% 
  pull(year)
highest_deaths_pct <- top_country_summary %>% 
  filter(metric == "Deaths") %>% 
  pull(percentage_difference)
list(
  value = paste0(highest_deaths_country_overall$location[1], " - ", format(highest_deaths_country_overall$max_deaths, big.mark = ","), " (", highest_deaths_year, ")")
)
```

## Global ROW 2 - Measles {height="1000px"}

### Global Measles maps {.tabset}

```{r}
#| title: Global Measles Cases

# 🚀 **Filtering the dataset to remove Antarctica & blank regions**
filtered_global_cases <- global_cases_deaths_summary %>%
  filter(location != "Antarctica" & total_cases > 0)  # ✅ Remove Antarctica & exclude empty data

# 🌍 Custom Color Gradient for Measles Cases
global_cases_colors <- list(
  c(0, "grey"),       # 0 cases
  c(0.1, "lightorange"),  
  c(0.3, "orange"),
  c(0.5, "red"),
  c(0.7, "darkred"),
  c(0.9, "maroon"),
  c(1, "black")  # Extreme cases (10M+)
)

# 🚀 Apply the corrected colorscale inside `plot_geo()`
global_map_cases <- plot_geo(filtered_global_cases, locationmode = "country names", frame = ~year) %>%
  add_trace(
    locations = ~location,
    z = ~total_cases,
    zmin = 0,
    zmax = 10000000,  # ✅ Set max value for color scale
    color = ~total_cases,
    colorscale = global_cases_colors,  # ✅ Now mapped directly to case numbers
    hoverinfo = "text",
    text = ~paste(location, "<br>Total Cases:", format(total_cases, big.mark = ",")), # CHANGED: Added format with comma separator
    marker = list(
      line = list(color = "black", width = 0.8)),
    colorbar = list(
      title = "Cases (Millions)",
      titleside = "bottom",
      titlefont = list(size = 16),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = 0.05,
      yanchor = "top",
      tickmode = "array",
      tickvals = c(0, 2000000, 4000000, 6000000, 8000000, 10000000),
      ticktext = c("0", "2", "4", "6", "8", "10"),
      len = 0.8,
      thickness = 15
    )) %>%
  layout(
    geo = list(
      showframe = FALSE,
      showcoastlines = TRUE,     # ✅ Enables coastlines
      showcountries = TRUE,      # ✅ Displays country borders
      showland = TRUE,           # ✅ Highlights land areas distinctly
      landcolor = "white",       # ✅ Keeps land visible while ensuring borders stand out
      countrycolor = "black",    # ✅ Ensures country borders appear in black
      projection = list(type = "mercator"),
      lataxis = list(range = c(-40, 75)),
      dragmode = FALSE,          # ✅ Disables dragging
      scrollZoom = FALSE         # ✅ Disables scroll zoom on geo plots
    ),
    margin = list(l = 30, r = 30, t = 50, b = 30)  # Standard margins
  ) %>%
  config(
    displayModeBar = FALSE,      # ✅ Removes the entire toolbar
    scrollZoom = FALSE,          # ✅ Disables scroll zooming
    doubleClick = FALSE,         # ✅ Disables double-click zoom
    staticPlot = FALSE           # ✅ Keeps interactivity but removes zoom
  )

global_map_cases
```

```{r}
#| title: Deaths

# Filter global data for deaths
filtered_global_deaths <- global_cases_deaths_summary %>%
  filter(location != "Antarctica" & total_cases > 0)  # ✅ Remove Antarctica & exclude empty data

# 🌍 Custom Color Gradient for Measles Deaths
global_deaths_colors <- list(
  c(0, "grey"),        # 0 deaths
  c(0.1, "lightyellow"),
  c(0.3, "lightorange"),  
  c(0.5, "orange"),
  c(0.7, "red"),
  c(0.9, "darkred"),
  c(1, "black")     # Extreme cases (500K+ deaths)
)

global_map_deaths <- plot_geo(filtered_global_deaths, 
                              locationmode = "country names", 
                              frame = ~year) %>%
  add_trace(
    locations = ~location,
    z = ~total_deaths,
    zmin = 0,
    zmax = 500000,  # ✅ Set max value for color scale
    color = ~total_deaths,
    colorscale = global_deaths_colors, 
    hoverinfo = "text",
    text = ~paste(location, "<br>Total Deaths:", format(total_deaths, big.mark = ",")), # CHANGED: Added format with comma separator
    marker = list(
      line = list(color = "black", width = 0.8)),
    colorbar = list(
      title = "Deaths (Thousands)",
      titleside = "bottom",
      titlefont = list(size = 16),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = 0.05,
      yanchor = "top",
      tickmode = "array",
      tickvals = c(0, 100000, 200000, 300000, 400000, 500000),
      ticktext = c("0", "100", "200", "300", "400", "500"),
      len = 0.8,
      thickness = 15
    )) %>%
  layout(
    geo = list(
      showframe = FALSE,
      showcoastlines = TRUE,     # ✅ Enables coastlines
      showcountries = TRUE,      # ✅ Displays country borders
      showland = TRUE,           # ✅ Highlights land areas distinctly
      landcolor = "white",       # ✅ Keeps land visible while ensuring borders stand out
      countrycolor = "black",    # ✅ Ensures country borders appear in black
      projection = list(type = "mercator"),
      lataxis = list(range = c(-40, 75)),
      dragmode = FALSE,          # ✅ Disables dragging
      scrollZoom = FALSE         # ✅ Disables scroll zoom on geo plots
    ),
    margin = list(l = 30, r = 30, t = 50, b = 30)  # Standard margins
  ) %>%
  config(
    displayModeBar = FALSE,      # ✅ Removes the entire toolbar
    scrollZoom = FALSE,          # ✅ Disables scroll zooming
    doubleClick = FALSE,         # ✅ Disables double-click zoom
    staticPlot = FALSE           # ✅ Keeps interactivity but removes zoom
  )

global_map_deaths
```

## GLOBAL ROW 3 - Continent boxes {height="250px"}

```{r}
#| include: false
#| label: continent_analysis
#| message: false

# Filter data based on selected continents
filtered_data <- country_cases_data %>%
  filter(continent %in% continents_available)

# Create continent statistics data for value boxes
continent_stats_data <- filtered_data %>%
  group_by(continent) %>%
  summarise(
    # Total cases and deaths for the continent
    total_cases = sum(total_cases, na.rm = TRUE),
    total_deaths = sum(total_deaths, na.rm = TRUE),
    
    # Average cases and deaths per country-year
    avg_cases_per_record = mean(total_cases, na.rm = TRUE),
    avg_deaths_per_record = mean(total_deaths, na.rm = TRUE),
    
    # Number of countries and records
    countries_count = n_distinct(location),
    total_records = n(),
    
    .groups = 'drop'
  ) %>%
  arrange(desc(total_cases))

# Find the country with highest cases in each continent
continent_highest_cases <- filtered_data %>%
  group_by(continent) %>%
  slice_max(total_cases, n = 1, with_ties = FALSE) %>%
  select(continent, location, total_cases, total_deaths, year) %>%
  rename(
    highest_cases_country = location,
    highest_cases_value = total_cases,
    highest_cases_deaths = total_deaths,
    highest_cases_year = year
  )

# Find the country with highest deaths in each continent
continent_highest_deaths <- filtered_data %>%
  group_by(continent) %>%
  slice_max(total_deaths, n = 1, with_ties = FALSE) %>%
  select(continent, location, total_cases, total_deaths, year) %>%
  rename(
    highest_deaths_country = location,
    highest_deaths_value = total_deaths,
    highest_deaths_cases = total_cases,
    highest_deaths_year = year
  )

# Combine all continent statistics
continent_stats_data <- continent_stats_data %>%
  left_join(continent_highest_cases, by = "continent") %>%
  left_join(continent_highest_deaths, by = "continent")

# Additional analysis: Top 3 countries per continent by cases
top_countries_per_continent <- filtered_data %>%
  group_by(continent) %>%
  slice_max(total_cases, n = 3, with_ties = FALSE) %>%
  select(continent, location, total_cases, total_deaths, year) %>%
  arrange(continent, desc(total_cases))

# Continent comparison metrics
continent_comparison <- continent_stats_data %>%
  select(continent, total_cases, total_deaths, countries_count, 
         highest_cases_country, highest_cases_value,
         highest_deaths_country, highest_deaths_value) %>%
  mutate(
    cases_per_country = round(total_cases / countries_count, 0),
    deaths_per_country = round(total_deaths / countries_count, 0),
    case_fatality_rate = round((total_deaths / total_cases) * 100, 2)
  )

# Year-over-year trends by continent
continent_yearly_trends <- filtered_data %>%
  group_by(continent, year) %>%
  summarise(
    yearly_cases = sum(total_cases, na.rm = TRUE),
    yearly_deaths = sum(total_deaths, na.rm = TRUE),
    countries_reporting = n_distinct(location),
    .groups = 'drop'
  ) %>%
  arrange(continent, year)

# Calculate percentage change for each continent (earliest vs latest year)
continent_change_analysis <- continent_yearly_trends %>%
  group_by(continent) %>%
  summarise(
    earliest_year = min(year),
    latest_year = max(year),
    earliest_cases = first(yearly_cases),
    latest_cases = last(yearly_cases),
    earliest_deaths = first(yearly_deaths),
    latest_deaths = last(yearly_deaths),
    .groups = 'drop'
  ) %>%
  mutate(
    cases_pct_change = round(((latest_cases - earliest_cases) / earliest_cases) * 100, 1),
    deaths_pct_change = round(((latest_deaths - earliest_deaths) / earliest_deaths) * 100, 1)
  )

# Final continent stats data with change metrics
continent_stats_data <- continent_stats_data %>%
  left_join(continent_change_analysis %>% 
            select(continent, cases_pct_change, deaths_pct_change), 
            by = "continent")

# Display summary
print("Continent Statistics Summary:")
print(continent_stats_data %>% 
      select(continent, total_cases, total_deaths, highest_cases_country, 
             highest_cases_value, cases_pct_change))
```

### Europe

```{r}
#| content: valuebox
#| title: "Europe"
#| icon: globe-europe-africa
#| color: "#2535D9"
#| fig-height: 250
europe_stats <- continent_stats_data %>% filter(continent == "Europe")
list(value = paste0(format(europe_stats$total_cases, big.mark = ",")))
```

### Americas

```{r}
#| content: valuebox
#| title: "Americas"
#| icon: globe-americas
#| color: "#B22222"
#| fig-height: 250
americas_stats <- continent_stats_data %>% filter(continent == "Americas")
list(value = paste0(format(americas_stats$total_cases, big.mark = ",")))
```

### Africa

```{r}
#| content: valuebox
#| title: "Africa"
#| icon: globe-europe-africa
#| color: "#65463E"
#| fig-height: 250
africa_stats <- continent_stats_data %>% filter(continent == "Africa")
list(value = paste0(format(africa_stats$total_cases, big.mark = ",")))
```

### Asia

```{r}
#| content: valuebox
#| title: "Asia"
#| icon: globe-central-south-asia
#| color: "#0A7029"
#| fig-height: 250
asia_stats <- continent_stats_data %>% filter(continent == "Asia")
list(value = paste0(format(asia_stats$total_cases, big.mark = ",")))
```

### Oceania

```{r}
#| content: valuebox
#| title: "Oceania"
#| icon: globe-asia-australia
#| color: "#C08D2C"
#| fig-height: 250
oceania_stats <- continent_stats_data %>% filter(continent == "Oceania")
list(value = paste0(format(oceania_stats$total_cases, big.mark = ",")))
```

## Global ROW 4 - Cases by Continent {height="800px"}

### Continent Cases {.tabset}

#### Continent Cases (bar view)

```{r}
# Prepare continental data
continent_cases <- filtered_global_cases %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent)) %>%
  group_by(continent, year) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(cases_millions = total_cases / 1e6)

continent_deaths <- filtered_global_deaths %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent)) %>%
  group_by(continent, year) %>%
  summarise(total_deaths = sum(total_deaths, na.rm = TRUE), .groups = "drop") %>%
  mutate(deaths_thousands = total_deaths / 1e3)

# Prepare country data with continents
country_cases_data <- filtered_global_cases %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent))

country_deaths_data <- filtered_global_deaths %>%
  left_join(continent_mapping, by = "location") %>%
  filter(!is.na(continent))

# Note: Y-axis ranges will adjust dynamically with the animation frames

# Continental Cases Bar Chart
continental_cases_bar <- continent_cases %>%
  plot_ly(
    x = ~continent,
    y = ~cases_millions,
    frame = ~year,
    type = 'bar',
    color = ~continent,
    colors = continent_colors,
    hovertemplate = paste(
      "<b>%{x}</b><br>",
      "Cases: %{y:.2f}M<br>",
      "Year: %{frame}<br>",
      "<extra></extra>"
    )
  ) %>%
  layout(
    xaxis = list(title = "Continent", titlefont = list(size = 14)),
    yaxis = list(
      title = "Cases (Millions)", 
      titlefont = list(size = 14),
      fixedrange = FALSE,
      autorange = TRUE
    ),
    showlegend = FALSE
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE,
    staticPlot = FALSE
  ) %>%
  animation_opts(frame = 1000, transition = 500) %>%
  animation_slider(currentvalue = list(prefix = "Year: "))

continental_cases_bar
```

#### (line view)

```{r}
# Continental Cases Line Chart ----
continental_cases_line <- continent_cases %>%
  plot_ly(
    x = ~year,
    y = ~cases_millions,
    color = ~continent,
    colors = continent_colors,
    type = 'scatter',
    mode = 'lines+markers',
    line = list(width = 3),
    marker = list(size = 6),
    hovertemplate = paste(
      "<b>%{fullData.name}</b><br>",
      "Year: %{x}<br>",
      "Cases: %{y:.2f}M<br>",
      "<extra></extra>"
    )
  ) %>%
  layout(
    xaxis = list(title = "Year", titlefont = list(size = 16)),
    yaxis = list(
      title = "Cases (Millions)", 
      titlefont = list(size = 16),
      fixedrange = FALSE,
      autorange = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = 1.02,
      yanchor = "bottom"
    )
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE,
    staticPlot = FALSE
  )

continental_cases_line
```

### Continental Deaths {.tabset}

#### Continental Deaths (Bar view)

```{r}
# Continental Deaths Bar Chart ----
continental_deaths_bar <- continent_deaths %>%
  plot_ly(
    x = ~continent,
    y = ~deaths_thousands,
    frame = ~year,
    type = 'bar',
    color = ~continent,
    colors = continent_colors,
    hovertemplate = paste(
      "<b>%{x}</b><br>",
      "Deaths: %{y:.1f}K<br>",
      "<extra></extra>"
    )
  ) %>%
  layout(
    xaxis = list(title = "Continent", titlefont = list(size = 14)),
    yaxis = list(
      title = "Deaths (Thousands)", 
      titlefont = list(size = 14),
      fixedrange = FALSE,
      autorange = TRUE
    ),
    showlegend = FALSE
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE,
    staticPlot = FALSE
  ) %>%
  animation_opts(frame = 1000, transition = 500) %>%
  animation_slider(currentvalue = list(prefix = "Year: "))

continental_deaths_bar
```

#### (Line view)

```{r}
continental_deaths_line <- continent_deaths %>%
  plot_ly(
    x = ~year,
    y = ~deaths_thousands,
    color = ~continent,
    colors = continent_colors,
    type = 'scatter',
    mode = 'lines+markers',
    line = list(width = 3),
    marker = list(size = 6),
    hovertemplate = paste(
      "<b>%{fullData.name}</b><br>",
      "Year: %{x}<br>",
      "Deaths: %{y:.1f}K<br>",
      "<extra></extra>"
    )
  ) %>%
  layout(
    xaxis = list(title = "Year", titlefont = list(size = 16)),
    yaxis = list(
      title = "Deaths (Thousands)", 
      titlefont = list(size = 16),
      fixedrange = FALSE,
      autorange = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = 1.02,
      yanchor = "bottom"
    )
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE,
    staticPlot = FALSE
  )

continental_deaths_line
```

## GLOBAL ROW 5 - MMR Vaccination Boxes {height="200px"}

### Average 1st Dose Coverage

```{r}
# Define the 4-shade purple color gradient
vaccination_colors_purple <- list(
  c(0, "#D3D3D3"),   # Grey (0-25%)
  c(0.25, "#9370DB"), # Medium Purple (25-50%)
  c(0.5, "#6A0DAD"),  # Dark Purple (50-75%)
  c(1, "#4B0082")    # Deep Dark Purple (75-100%)
)

# Read and process first dose data
vac_1dose <- read_csv("data/measles_global/1dose_share-of-children-vaccinated-against-measles.csv") %>%
  rename(`1_dose` = `Share of one-year-olds who have received their first dose of measles-containing vaccine (MCV1)`,
         `location` = `Entity`, `year` = `Year`) %>%
  left_join(country_coords, by = c("location" = "country"))

# Read and process second dose data
vac_2dose <- read_csv("data/measles_global/2_doses_ share-of-children-vaccinated-with-mcv.csv") %>%
  rename(`2_dose` = `Share of children who have received two doses of measles-containing vaccine as per the national schedule (MCV2)`,
         `location` = `Entity`, `year` = `Year`) %>%
  left_join(country_coords, by = c("location" = "country"))

# Identify the later start year
start_year_1dose <- min(vac_1dose$year, na.rm = TRUE)
start_year_2dose <- min(vac_2dose$year, na.rm = TRUE)
later_start_year <- max(start_year_1dose, start_year_2dose)

# Filter datasets to match the later start year
vac_1dose_filtered <- vac_1dose %>% filter(year >= later_start_year)
vac_2dose_filtered <- vac_2dose %>% filter(year >= later_start_year)

# Merge both datasets
vac_combined <- vac_1dose_filtered %>%
  left_join(vac_2dose_filtered %>% select(location, year, `2_dose`), by = c("location", "year"))

vac_combined <- vac_combined %>%
  rename(dose_1 = `1_dose`,
         dose_2 = `2_dose`)
```

### 1st Dose Global Coverage

```{r}
#| content: valuebox
#| title: "1st Dose Global Coverage"
#| icon: shield-check
#| color: purple

list(
  value = paste0(format(round(mean(vac_combined$dose_1, na.rm = TRUE), 0)), "%")
)
```

### Average 2nd Dose Coverage

```{r}
#| content: valuebox
#| title: "2nd Dose Global Coverage"
#| icon: shield-plus
#| color: purple

list(
  value = paste0(format(round(mean(vac_combined$dose_2, na.rm = TRUE), 0)), "%")
)
```

### Highest Vaccination Coverage

```{r}
#| content: valuebox
#| title: "Highest Coverage Country"
#| icon: shield-lock
#| color: purple

list(
  value = vac_combined %>%
    filter(dose_2 == max(dose_2, na.rm = TRUE)) %>%
    slice(1) %>%
    pull(location),
  caption = paste0(round(
    max(vac_combined$dose_2, na.rm = TRUE), 0
  ), "% coverage")
)
```

## GLOBAL ROW 6 - 🌍 Interactive Vaccination Coverage Maps {height="1000px"}

### Vaccination coverage {.tabset}

#### Global Trends in MMR Vaccination Coverage Over Time (1st Dose)

```{r}
# 🚀 **Filtering the dataset to remove Antarctica & blank regions**
filtered_vac1 <- vac_combined

vaccination_map_1st <- plot_geo(filtered_vac1, 
                                locationmode = "country names", 
                                frame = ~year) %>%
  add_trace(
    locations = ~location,
    z = ~dose_1,
    zmin = 0,
    zmax = 100,
    color = ~dose_1,
    colorscale = vaccination_colors_purple,
    hoverinfo = "text",
    text = ~paste(location, "<br>Year:", year, "<br>1st Dose Coverage:", round(dose_1, 1), "%"),
    marker = list(line = list(color = "black", width = 0.8))
  ) %>%
  colorbar(
    title = list(
      text = "%",
      side = "left",           
      font = list(size = 14)  
    ),
    orientation = "v",          # Vertical orientation
    x = 1.02,                   # Position to the right of the plot
    y = 0.75,                   # Position in top area
    len = 0.6,                  # Increased length (60% of plot height)
    thickness = 15,             # Keeps the colorbar thin
    xanchor = "left",           # Anchors from the left side
    yanchor = "top"             # Anchors from the top
  ) %>%
  layout(
    geo = list(
      showframe = FALSE,
      showcoastlines = TRUE,
      showcountries = TRUE,
      showland = TRUE,
      landcolor = "white",
      countrycolor = "black",
      projection = list(type = "mercator"),
      lataxis = list(range = c(-40, 75)),
      dragmode = FALSE,
      scrollZoom = FALSE
    ),
    margin = list(l = 50, r = 120, t = 50, b = 50)  # Increased right margin for colorbar + title
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE,
    staticPlot = FALSE
  )

vaccination_map_1st
```

#### 2nd Dose (Fully vaccininated)

```{r}
# Vaccination 2 dose map ----
filtered_vac2 <- vac_combined 

vaccination_map_2nd <- plot_geo(filtered_vac2, 
                                locationmode = "country names", 
                                frame = ~year) %>%
  add_trace(
    locations = ~location,
    z = ~ dose_2,
    zmin = 0,
    zmax = 100,
    color = ~ dose_2,
    colorscale = vaccination_colors_purple,
    hoverinfo = "text",
    text = ~paste(location, "<br>Year:", year, "<br>1st Dose Coverage:", round(dose_2, 1), "%"),
    marker = list(line = list(color = "black", width = 0.8))
  ) %>%
  colorbar(
    title = list(
      text = "%",
      side = "left",           
      font = list(size = 14)  
    ),
    orientation = "v",          # Vertical orientation
    x = 1.02,                   # Position to the right of the plot
    y = 0.75,                   # Position in top area
    len = 0.6,                  # Increased length (60% of plot height)
    thickness = 15,             # Keeps the colorbar thin
    xanchor = "left",           # Anchors from the left side
    yanchor = "top"             # Anchors from the top
  ) %>%
  layout(
    geo = list(
      showframe = FALSE,
      showcoastlines = TRUE,
      showcountries = TRUE,
      showland = TRUE,
      landcolor = "white",
      countrycolor = "black",
      projection = list(type = "mercator"),
      lataxis = list(range = c(-40, 75)),
      dragmode = FALSE,
      scrollZoom = FALSE
    ),
    margin = list(l = 50, r = 120, t = 50, b = 50)  # Increased right margin for colorbar + title
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE,
    staticPlot = FALSE
  )

vaccination_map_2nd
```

## GLOBAL ROW 7 - SOURCES {height="400px"}

### 🌍 Global Data Sources

**Global Cases and Deaths**

-   **File**: `global_cases_deaths.csv`

-   **Source**: Global Burden of Disease Study 2021 (GBD 2021) Results

-   **Institution**: Institute for Health Metrics and Evaluation (IHME), Seattle, United States

-   **Year**: 2022

-   **Citation**: Global Burden of Disease Collaborative Network. Global Burden of Disease Study 2021 (GBD 2021) Results. Seattle, United States: Institute for Health Metrics and Evaluation (IHME), 2022. Available from https://vizhub.healthdata.org/gbd-results/

**Global Cases and Deaths by Demographics**

-   **File**: `global_cases_sex_age.csv`

-   **Source**: Global Burden of Disease Study 2021 (GBD 2021) Results

-   **Institution**: Institute for Health Metrics and Evaluation (IHME), Seattle, United States

-   **Year**: 2022

-   **Citation**: Global Burden of Disease Collaborative Network. Global Burden of Disease Study 2021 (GBD 2021) Results. Seattle, United States: Institute for Health Metrics and Evaluation (IHME), 2022. Available from https://vizhub.healthdata.org/gbd-results/

**Global Vaccination Coverage - First Dose**

-   **File**: `1dose_share-of-children-vaccinated-against-measles.csv`

-   **Source**: [Share of one-year-olds vaccinated against measles, 2023](https://ourworldindata.org/grapher/share-of-children-vaccinated-against-measles)

-   **Institution**: Our World in Data

-   **Coverage**: Global vaccination rates for measles first dose (MCV1)

**Global Vaccination Coverage - Second Dose**

-   **File**: `2_doses_share-of-children-vaccinated-with-mcv.csv`

-   **Source**: [Share of children fully vaccinated against measles, 2023](https://ourworldindata.org/grapher/share-of-children-vaccinated-with-mcv2)

-   **Institution**: Our World in Data

-   **Coverage**: Global vaccination rates for the second dose (MCV2)

# UK

## UK ROW 1 - UK BOXES {height="250px"}

```{r}
#| include: false
#| label: uk_boxes_first
#| message: false

library(scales)

region_coords <- tibble(
  region = c("North_East", "North_West", "Yorkshire_and_the_Humber",
             "East_Midlands", "West_Midlands", "East_of_England",
             "London", "South_East", "South_West", "Wales"),
  lat = c(54.97, 53.48, 53.80, 52.95, 52.48, 52.37, 51.50, 51.27, 50.85, 52.30),
  lon = c(-1.61, -2.24, -1.54, -1.15, -1.90, 0.00, -0.12, -0.47, -3.60, -3.78))

# Cases
uk_cases <- read_xlsx("data/measles_uk/cases_region_2012_2024.xlsx") %>%
  pivot_longer( # Pivot longer to get regions and cases in one column
    cols = c("North_East", "North_West", "Yorkshire_and_the_Humber",
             "East_Midlands", "West_Midlands", "East_of_England",
             "London", "South_East", "South_West", "Wales"),
    names_to = "region",
    values_to = "cases") %>%
  left_join(region_coords, by = "region")

# Sum cases by year, region and age group
uk_cases_summary1 <- uk_cases %>%
  group_by(Year, region, Age_group) %>% # Group by both region and age group
  summarize(total_cases = sum(cases, na.rm = TRUE)) %>%
  ungroup()

uk_cases_summary1 <- uk_cases_summary1 %>%
  mutate(
    age = case_when(
      Age_group == "less than 1 year" ~ "<1",
      Age_group == "1 to 4 years" ~ "1-4",
      Age_group == "5 to 9 years" ~ "5-9",
      Age_group == "10 to 14 years" ~ "10-14",
      Age_group == "15 to 19 years" ~ "15-19",
      Age_group == "20 to 24 years" ~ "20-24",
      Age_group == "25 to 29 years" ~ "25-29",
      Age_group == "30 to 34 years" ~ "30-34",
      Age_group == "35 years and older" ~ "35+",
      TRUE ~ "no_data"  # Explicitly marking unknown values
    )
  )

# Load UK region boundaries
uk_regions <- st_read("data/measles_uk/rgn2025.geojson")

# Create region mapping
region_mapping <- data.frame(
  data_region = c("North_East", "North_West", "Yorkshire_and_the_Humber",
                  "East_Midlands", "West_Midlands", "East_of_England",
                  "London", "South_East", "South_West", "Wales", "Scotland"),
  geo_region = c("North East", "North West", "Yorkshire and The Humber",
                 "East Midlands", "West Midlands", "East of England",
                 "London", "South East", "South West", "Wales", "Scotland")
)

# Prepare cases data with proper region names
uk_cases_age_clean <- uk_cases_summary1 %>%
  left_join(region_mapping, by = c("region" = "data_region")) %>%
  select(-region) %>%
  rename(region = geo_region) %>%
  mutate(
    total_cases = as.numeric(total_cases),
    Year = as.numeric(Year)
  ) %>%
  filter(!is.na(region))

# Rename columns and select relevant data
filtered_age_cases <- uk_cases_age_clean %>%
  select(Year, region, Age_group, total_cases) %>%
  rename(
    Year = Year,
    `Age Group` = Age_group,
    Region = region,
    Cases = total_cases
  )

# Calculate total cases across all years and regions
total_uk_cases <- sum(uk_cases_age_clean$total_cases, na.rm = TRUE)

# Calculate year-over-year change (2023 vs 2024 or latest available years)
latest_years <- sort(unique(uk_cases_age_clean$Year), decreasing = TRUE)[1:2]
latest_year_cases <- uk_cases_age_clean %>%
  filter(Year == latest_years[1]) %>%
  summarise(total = sum(total_cases, na.rm = TRUE)) %>%
  pull(total)

previous_year_cases <- uk_cases_age_clean %>%
  filter(Year == latest_years[2]) %>%
  summarise(total = sum(total_cases, na.rm = TRUE)) %>%
  pull(total)

cases_pct_change <- ((latest_year_cases - previous_year_cases) / previous_year_cases) * 100

# Find highest cases by region
highest_region_cases <- uk_cases_age_clean %>%
  group_by(region) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
  arrange(desc(total_cases)) %>%
  slice(1)

# Find the year when this region had its highest cases
highest_region_year <- uk_cases_age_clean %>%
  filter(region == highest_region_cases$region) %>%
  group_by(Year) %>%
  summarise(yearly_total = sum(total_cases, na.rm = TRUE)) %>%
  arrange(desc(yearly_total)) %>%
  slice(1) %>%
  pull(Year)

# Find highest cases by age group
highest_age_cases <- uk_cases_age_clean %>%
  group_by(Age_group) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
  arrange(desc(total_cases)) %>%
  slice(1)

# Find the year when this age group had its highest cases
highest_age_year <- uk_cases_age_clean %>%
  filter(Age_group == highest_age_cases$Age_group) %>%
  group_by(Year) %>%
  summarise(yearly_total = sum(total_cases, na.rm = TRUE)) %>%
  arrange(desc(yearly_total)) %>%
  slice(1) %>%
  pull(Year)

# Calculate children under 5 percentage
children_under_5_cases <- uk_cases_age_clean %>%
  filter(Age_group %in% c("less than 1 year", "1 to 4 years")) %>%
  summarise(total = sum(total_cases, na.rm = TRUE)) %>%
  pull(total)

children_under_5_pct <- round((children_under_5_cases / total_uk_cases) * 100)

# Calculate average annual cases
avg_annual_cases <- round(total_uk_cases / length(unique(uk_cases_age_clean$Year)))

# Calculate peak year data
peak_year_data <- uk_cases_age_clean %>%
  group_by(Year) %>%
  summarise(yearly_total = sum(total_cases, na.rm = TRUE)) %>%
  arrange(desc(yearly_total)) %>%
  slice(1)
```

### Value Box a

```{r}
#| content: valuebox
#| title: "Total Measles Cases"
#| icon: virus
#| color: danger
list(
  value = paste0(format(total_uk_cases, big.mark = ","), " (", round(cases_pct_change, 0), "% change)")
)
```

### Value Box b

```{r}
#| content: valuebox
#| title: "Highest Recorded Cases by Region"
#| icon: virus
#| color: danger
list(
  value = paste0(highest_region_cases$region, " - ", format(highest_region_cases$total_cases, big.mark = ","), " (", highest_region_year, ")")
)
```

### Value Box c

```{r}
#| content: valuebox
#| title: "Highest Recorded Cases By Age Group"
#| icon: person-fill
#| color: info
list(
  value = paste0(highest_age_cases$Age_group, " - ", format(highest_age_cases$total_cases, big.mark = ","), " (", highest_age_year, ")")
)
```

### Value Box d

```{r}
#| content: valuebox
#| title: "Most Vulnerable Population"
#| icon: person-fill
#| color: info
list(
  value = paste0("Children Under 5 - ", children_under_5_pct, "% of all cases")
)
```

## UK ROW 3 - REGION TABLE PLOTS {height="1000px"}

### UK region age cases column {width="60%"}

#### UK Measles Cases by Region (bar) {.tabset}

##### UK Measles Cases by Region (Bar view)

```{r}
# Load required libraries
library(RColorBrewer)
library(crosstalk)

# Define the RColorBrewer Paired palette
color_palette <- brewer.pal(10, "Paired")  # Adjust number based on unique groups

# 3. ANIMATED BAR CHART BY YEAR - UK Measles Cases by Region
uk_cases_region_bar <- filtered_age_cases %>%
  group_by(Year, Region) %>%
  summarise(total_cases = sum(Cases, na.rm = TRUE), .groups = "drop") %>%
  plot_ly(y = ~reorder(Region, total_cases), x = ~total_cases, 
          frame = ~Year, 
          type = 'bar',
          color = ~Region, 
          colors = color_palette,
          orientation = 'h',
          hovertemplate = paste('<b>%{y}</b><br>',
                                'Total Cases: %{x}<br>',
                                '<extra></extra>')) %>%
  layout(xaxis = list(title = "Total Cases"),
         yaxis = list(title = ""),
         showlegend = FALSE) %>%
  animation_opts(frame = 300, transition = 200) %>%
  config(displayModeBar = FALSE)

uk_cases_region_bar
```

##### (Line view)

```{r}

# Define the RColorBrewer Paired palette
color_palette <- brewer.pal(12, "Paired")  # Adjust number based on unique groups

# Function to create animated line plot with moving dots, incorporating custom color palette
create_animated_plot <- function(data, breakdown_type = "region") {
  
  if (breakdown_type == "age") {
    plot_data <- data %>%
      mutate(group_var = case_when(
        `Age Group` == "less than 1 year" ~ "less than 1",
        `Age Group` == "1 to 4 years" ~ "1-4",
        `Age Group` == "5 to 9 years" ~ "5-9",
        `Age Group` == "10 to 14 years" ~ "10-14",
        `Age Group` == "15 to 19 years" ~ "15-19",
        `Age Group` == "20 to 24 years" ~ "20-24",
        `Age Group` == "25 to 29 years" ~ "25-29",
        `Age Group` == "30 to 34 years" ~ "30-34",
        `Age Group` == "35 years and older" ~ "35+",
        TRUE ~ `Age Group`
      )) %>%
      mutate(group_var = factor(group_var, 
                                levels = c("less than 1", "1-4", "5-9", "10-14", "15-19", 
                                           "20-24", "25-29", "30-34", "35+"), 
                                ordered = TRUE)) %>%
      arrange(group_var) %>%
      group_by(Year, group_var) %>%
      summarise(Cases = sum(Cases, na.rm = TRUE), .groups = "drop")
    
    # Age group legend with explicit ordering
    legend_config <- list(
      orientation = "h",
      x = 0.5,
      xanchor = 'center',
      y = 1.02,
      yanchor = 'bottom',
      bgcolor = 'rgba(255, 255, 255, 0.9)',
      bordercolor = 'rgba(0, 0, 0, 0.3)',
      borderwidth = 1,
      traceorder = "normal",  # This helps maintain order
      font = list(size = 14)
    )
    
  } else {
    plot_data <- data %>%
      mutate(group_var = Region) %>%
      group_by(Year, group_var) %>%
      summarise(Cases = sum(Cases, na.rm = TRUE), .groups = "drop")
    
    legend_config <- list(
      orientation = "h",
      x = 0.5,
      xanchor = 'center',
      y = 1.02,
      yanchor = 'bottom',
      bgcolor = 'rgba(255, 255, 255, 0.9)',
      bordercolor = 'rgba(0, 0, 0, 0.3)',
      borderwidth = 1,
      itemsizing = "constant",
      itemwidth = 20,
      font = list(size = 14)
    )
  }
  
  years <- sort(unique(plot_data$Year))
  groups <- unique(plot_data$group_var)
  
  # For age plots, ensure groups are in the correct order
  if (breakdown_type == "age") {
    age_order <- c("less than 1", "1-4", "5-9", "10-14", "15-19", 
                   "20-24", "25-29", "30-34", "35+")
    groups <- age_order[age_order %in% groups]  # Keep only groups that exist in data
  }
  
  # Create animated data
  animated_data <- data.frame()
  
  for(current_year in years) {
    frame_data <- plot_data %>%
      filter(Year <= current_year) %>%
      mutate(frame = current_year)
    
    animated_data <- rbind(animated_data, frame_data)
  }
  
  # Create the plot
  p <- plot_ly()
  
  # Add traces in the correct order for age plots
  if (breakdown_type == "age") {
    for(i in seq_along(groups)) {
      group_data <- animated_data %>% filter(group_var == groups[i])
      
      p <- p %>%
        add_trace(
          data = group_data,
          x = ~Year,
          y = ~Cases,
          name = groups[i],
          color = I(color_palette[i]),  # Use specific color from palette
          frame = ~frame,
          type = 'scatter',
          mode = 'lines+markers',
          line = list(width = 3),
          marker = list(size = 8),
          hovertemplate = paste(
            '<b>Year:</b> %{x}<br>',
            '<b>Cases:</b> %{y}<br>',
            '<b>', groups[i], '</b><br>',
            '<extra></extra>'
          )
        )
    }
  } else {
    # For region plots, use the original method
    p <- p %>%
      add_trace(
        data = animated_data,
        x = ~Year,
        y = ~Cases,
        color = ~group_var,
        colors = color_palette,
        frame = ~frame,
        type = 'scatter',
        mode = 'lines+markers',
        line = list(width = 3),
        marker = list(size = 8),
        hovertemplate = paste(
          '<b>Year:</b> %{x}<br>',
          '<b>Cases:</b> %{y}<br>',
          '<b>%{fullData.name}<br>',
          '<extra></extra>'
        )
      )
  }
  
  p <- p %>%
    layout(
      xaxis = list(
        title = "Year",
        titlefont = list(size = 16),
        range = c(min(years) - 0.5, max(years) + 0.5)
      ),
      yaxis = list(
        title = "Number of Cases",
        titlefont = list(size = 16),
        range = c(0, max(plot_data$Cases) * 1.1)
      ),
      plot_bgcolor = 'white',
      paper_bgcolor = 'white',
      legend = legend_config,
      margin = list(t = 30)  
    ) %>%
    animation_opts(
      frame = 500,
      transition = 300,
      redraw = FALSE
    ) %>%
    animation_slider(
      currentvalue = list(
        prefix = "Year: ",
        font = list(size = 16)
      )
    ) %>%
    animation_button(
      x = 0, xanchor = 'left', y = -0.1, yanchor = 'bottom'
    ) %>%
    config(displayModeBar = FALSE)
  
  return(p)
}

uk_cases_region_line <- create_animated_plot(filtered_age_cases, "region")

uk_cases_region_line
```

##### UK Measles Cases by Age Group (Bar view)

```{r}

# Define the RColorBrewer Paired palette
color_palette <- brewer.pal(12, "Paired")  # Adjust number based on unique groups

# Create animated bar chart by age group
uk_cases_age_bar <- filtered_age_cases %>%
  mutate(Age_Group_Short = case_when(
    `Age Group` == "less than 1 year" ~ "less than 1",
    `Age Group` == "1 to 4 years" ~ "1-4",
    `Age Group` == "5 to 9 years" ~ "5-9",
    `Age Group` == "10 to 14 years" ~ "10-14",
    `Age Group` == "15 to 19 years" ~ "15-19",
    `Age Group` == "20 to 24 years" ~ "20-24",
    `Age Group` == "25 to 29 years" ~ "25-29",
    `Age Group` == "30 to 34 years" ~ "30-34",
    `Age Group` == "35 years and older" ~ "35+",
    TRUE ~ `Age Group`
  )) %>%
  mutate(Age_Group_Short = factor(Age_Group_Short, 
                                  levels = c("less than 1", "1-4", "5-9", "10-14", "15-19", 
                                             "20-24", "25-29", "30-34", "35+"), 
                                  ordered = TRUE)) %>%
  group_by(Year, Age_Group_Short) %>%
  summarise(total_cases = sum(Cases, na.rm = TRUE), .groups = "drop") %>%
  plot_ly(y = ~Age_Group_Short, x = ~total_cases,  # Changed this line
          frame = ~Year, 
          type = 'bar',
          color = ~Age_Group_Short, 
          colors = color_palette,
          orientation = 'h',
          hovertemplate = paste('<b>%{y}</b><br>',
                                'Total Cases: %{x}<br>',
                                '<extra></extra>')) %>%
  layout(xaxis = list(title = "Total Cases"),
         yaxis = list(title = ""),
         showlegend = FALSE) %>%
  animation_opts(frame = 1000, transition = 1000) %>%
  config(displayModeBar = FALSE)

uk_cases_age_bar
```

##### (Line view)

```{r}

uk_cases_age_line <- create_animated_plot(filtered_age_cases, "age")

uk_cases_age_line
```

### UK Map column {width="40%"}

```{r}
#| title: Measles Cases Across UK Regions (2012-2024)

# Define UK region coordinates
region_coords <- tibble(
  region = c("North_East", "North_West", "Yorkshire_and_the_Humber",
             "East_Midlands", "West_Midlands", "East_of_England",
             "London", "South_East", "South_West", "Wales"),
  lat = c(54.97, 53.48, 53.80, 52.95, 52.48, 52.37, 51.50, 51.27, 50.85, 52.30),
  lon = c(-1.61, -2.24, -1.54, -1.15, -1.90, 0.00, -0.12, -0.47, -3.60, -3.78))

# Load UK measles case data
uk_cases <- read_xlsx("data/measles_uk/cases_region_2012_2024.xlsx") %>%
  pivot_longer(cols = region_coords$region, 
               names_to = "region", values_to = "cases") %>%
  left_join(region_coords, by = "region")

# Summarize cases by year, region, and age group
uk_cases_summary <- uk_cases %>%
  group_by(Year, region, Age_group) %>%
  summarize(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(age = case_when(
    Age_group == "less than 1 year" ~ "<1",
    Age_group == "1 to 4 years" ~ "1-4",
    Age_group == "5 to 9 years" ~ "5-9",
    Age_group == "10 to 14 years" ~ "10-14",
    Age_group == "15 to 19 years" ~ "15-19",
    Age_group == "20 to 24 years" ~ "20-24",
    Age_group == "25 to 29 years" ~ "25-29",
    Age_group == "30 to 34 years" ~ "30-34",
    Age_group == "35 years and older" ~ "35+",
    TRUE ~ "no_data"
  ))

# Load UK region boundaries without printing metadata
uk_regions <- st_read("data/measles_uk/rgn2025.geojson", quiet = TRUE)

# Define region mapping
region_mapping <- tibble(
  data_region = c(region_coords$region, "Scotland"),
  geo_region = c("North East", "North West", "Yorkshire and The Humber",
                 "East Midlands", "West Midlands", "East of England",
                 "London", "South East", "South West", "Wales", "Scotland")
)

# Prepare cases data with proper region names
uk_cases_clean <- uk_cases_summary %>%
  left_join(region_mapping, by = c("region" = "data_region")) %>%
  select(-region) %>%
  rename(region = geo_region) %>%
  mutate(total_cases = as.numeric(total_cases),
         Year = as.numeric(Year)) %>%
  filter(!is.na(region))

# Insert Scotland data with NA values
scotland_entry <- tibble(
  Year = unique(uk_cases_clean$Year),
  Age_group = "All Ages",
  region = "Scotland",
  total_cases = NA_real_
)

uk_cases_clean <- bind_rows(uk_cases_clean, scotland_entry)

# Final case dataset preparation
complete_cases <- uk_cases_clean %>%
  mutate(
    total_cases = case_when(
      region == "Scotland" ~ NA_real_,
      is.na(total_cases) ~ 0,
      TRUE ~ total_cases
    ),
    hover_text = case_when(
      region == "Scotland" ~ paste("Region:", region, "<br>Year:", Year, "<br>Cases: No data"),
      TRUE ~ paste("Region:", region, "<br>Year:", Year, "<br>Cases:", total_cases)
    )
  )

# Merge with geospatial data
cases_sf_timeline <- uk_regions %>%
  rename(region = areanm) %>%  # Rename `areanm` to match `complete_cases`
  left_join(complete_cases, by = "region") %>%
  filter(!is.na(Year))

# Define color scale limits
color_limits <- c(0, max(cases_sf_timeline$total_cases, na.rm = TRUE))

# Create animated map plot with legend positioned on right
# Calculate exact bbox from your data for optimal fitting
bbox <- st_bbox(cases_sf_timeline)

# Create animated map plot with auto-fitting dimensions
gg_animated <- ggplot(cases_sf_timeline) +
  geom_sf(aes(fill = total_cases, frame = Year, text = hover_text), 
          color = "black", linewidth = 0.2) +
  scale_fill_gradientn(
    colors = c("#E5C1F5", "#C490D1", "#944FB2", "#5B2271", "#3B1055"),
    na.value = "grey70",
    limits = color_limits,
    name = "Cases",
    guide = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      title.theme = element_text(size = 10),
      barwidth = 1,
      barheight = 8
    )
  ) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]), 
           ylim = c(bbox["ymin"], bbox["ymax"]), 
           expand = FALSE) +
  theme_void() +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "right",
    legend.margin = margin(0, 0, 0, 2),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank()
  )

# Convert to interactive Plotly visualization with zoom functionality enabled
uk_animated_map <- ggplotly(gg_animated, tooltip = "text") %>%
  animation_opts(frame = 500, transition = 300, redraw = TRUE) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Year: ", 
      font = list(color = "black", size = 10)
    ),
    x = 0.1, y = 0, xanchor = 'left', yanchor = 'top',
    len = 0.75,
    thickness = 12
  ) %>%
  animation_button(
    x = 0.88, xanchor = 'left', y = 0, yanchor = 'top',
    width = 50, height = 20
  ) %>%
  config(
    displayModeBar = TRUE,        # Enable mode bar for zoom tools
    modeBarButtonsToRemove = c(   # Remove unwanted tools, keep zoom
      'pan2d', 'select2d', 'lasso2d', 'autoScale2d', 
      'hoverClosestCartesian', 'hoverCompareCartesian',
      'toggleSpikelines', 'resetScale2d'
    ),
    scrollZoom = TRUE,            # Enable scroll wheel zoom
    doubleClick = 'reset',        # Double-click to reset zoom
    showTips = FALSE, 
    displaylogo = FALSE,
    responsive = TRUE
  ) %>%
  layout(
    autosize = TRUE,
    xaxis = list(
      fixedrange = FALSE,         # Allow zooming on x-axis
      showgrid = FALSE, 
      zeroline = FALSE, 
      showline = FALSE,
      showticklabels = FALSE
    ),
    yaxis = list(
      fixedrange = FALSE,         # Allow zooming on y-axis
      showgrid = FALSE, 
      zeroline = FALSE, 
      showline = FALSE,
      showticklabels = FALSE
    ),
    margin = list(t = 5, l = 5, r = 5, b = 25),
    showlegend = TRUE,
    dragmode = 'zoom'             # Set default interaction to zoom
  )

uk_animated_map
```

## UK ROW 4 - TABLE {height="600px"}

```{r}
#| title: "UK Measles Cases Across UK Regions and Age Groups"

# Define color scale for `Cases`
color_scale_uk_cases <- c("#F3E2FC", "#DDBDE6", "#B898D0")

# Generate proper breakpoints for coloring
breaks <- quantile(filtered_age_cases$Cases, probs = seq(0, 1, length.out = 3), na.rm = TRUE)

# Apply professional styling with colors added
uk_ages_cases_table <- datatable(filtered_age_cases, 
          rownames = FALSE,
          class = "compact stripe hover order-column",  
          options = list(
            pageLength = 10,   # Show only 10 rows
            lengthMenu = list(c(10, 25, 50, 100), c("10", "25", "50", "100")), 
            autoWidth = FALSE,  
            scrollY = FALSE,   # Disable vertical scrolling
            scrollX = FALSE,   # Disable horizontal scrolling
            paging = TRUE,     # Enable pagination
            dom = '<"top-left-search"f>rt<"bottom"lp><"clear">',
            buttons = c("csv", "excel"),
            columnDefs = list(
              list(targets = "_all", className = "dt-left")  
            ),
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'height': '450px'});", # Fixed container height
              "}"
            )
          ), 
          filter = "top") %>%
  formatStyle(
    columns = c("Year", "Age Group", "Region"),
    fontFamily = "Arial",
    fontSize = "13px",
    textAlign = "left"
  ) %>%
  formatStyle(
    "Cases",
    backgroundColor = styleInterval(
      breaks[-1],  # Remove first element to match expected length
      color_scale_uk_cases
    ),
    fontFamily = "Arial",
    fontSize = "13px",
    textAlign = "left"
  )

uk_ages_cases_table
```

## UK ROW 5 - VAC BOXES {height="200px"}

```{r}
#| include: false
#| label: uk_vac_calculations
#| message: false

uk_vac_completion <- read_xlsx("data/measles_uk/measles_vac_complete_england_2003_2024.xlsx") 
uk_vac_primary <- read_xlsx("data/measles_uk/measles_vac_primary_england_2003_2024.xlsx")

# Get the latest year's data
latest_complete <- uk_vac_completion %>% 
  filter(year == max(year)) %>% 
  pull(vac_complete)

latest_primary <- uk_vac_primary %>% 
  filter(year == max(year)) %>% 
  pull(vac_primary)

# Calculate coverage gap from WHO target
coverage_gap <- 95 - latest_complete  # WHO target is 95%

# Calculate years tracked
years_tracked <- length(unique(uk_vac_completion$year))

# Calculate trend over last 5 years
recent_data <- uk_vac_completion %>% 
  arrange(year) %>% 
  tail(5)

trend_direction <- ifelse(
  tail(recent_data$vac_complete, 1) > head(recent_data$vac_complete, 1),
  "↗ Improving",
  "↘ Declining"
)

# Calculate year-over-year change
latest_years <- sort(unique(uk_vac_completion$year), decreasing = TRUE)[1:2]
latest_year_complete <- uk_vac_completion %>%
  filter(year == latest_years[1]) %>%
  pull(vac_complete)

previous_year_complete <- uk_vac_completion %>%
  filter(year == latest_years[2]) %>%
  pull(vac_complete)

coverage_change <- latest_year_complete - previous_year_complete
```

### UK VAC Box a

```{r}
#| content: valuebox
#| title: "Complete Course Coverage"
#| icon: shield-check
#| color: primary
list(
  value = paste0(latest_complete, "%")
)
```

### UK VAC Box b

```{r}
#| content: valuebox
#| title: "Primary Course Coverage"
#| icon: hospital
#| color: success
list(
  value = paste0(latest_primary, "%")
)
```

### UK VAC Box c

```{r}
#| content: valuebox
#| title: "Gap to WHO Target (95%)"
#| icon: exclamation-triangle
#| color: danger
list(
  value = paste0(abs(coverage_gap), "% ", ifelse(coverage_gap > 0, "below", "above"))
)
```

### UK VAC Box d

```{r}
#| content: valuebox
#| title: "Years of Data Available"
#| icon: calendar
#| color: info
list(
  value = paste0(years_tracked, " years")
)
```

## UK ROW 6 - VAC PLOTS {height="800px"}

### UK vaccination coverage column 1 {.tabset}

#### UK Trends in MMR Vaccination Coverage Over Time (Bar view)

```{r}

uk_vac_completion <- read_xlsx("data/measles_uk/measles_vac_complete_england_2003_2024.xlsx") 
uk_vac_primary <- read_xlsx("data/measles_uk/measles_vac_primary_england_2003_2024.xlsx")

# Define purple color palette
purple_palette <- brewer.pal(9, "Purples")[c(5, 7)]  # Select two nice purple shades

# 1. ANIMATED LINE PLOT - Vaccination Coverage Over Time
# Combine both datasets for comparison
vac_combined <- uk_vac_completion %>%
  rename(coverage = vac_complete) %>%
  mutate(vaccination_type = "Complete Course") %>%
  bind_rows(
    uk_vac_primary %>%
      rename(coverage = vac_primary) %>%
      mutate(vaccination_type = "Primary Course")
  )

# Create animated data for cumulative lines
years <- sort(unique(vac_combined$year))
vac_types <- unique(vac_combined$vaccination_type)

animated_vac_data <- data.frame()

for(current_year in years) {
  frame_data <- vac_combined %>%
    filter(year <= current_year) %>%
    mutate(frame = current_year)
  
  animated_vac_data <- rbind(animated_vac_data, frame_data)
}


# 2. ANIMATED BAR CHART - UK MMR Vaccination Coverage
uk_vaccination_coverage_bar <- vac_combined %>%
  plot_ly(x = ~reorder(vaccination_type, coverage), y = ~coverage, 
          frame = ~year, 
          type = 'bar',
          color = ~vaccination_type, 
          colors = purple_palette,  # Use purple colors
          hovertemplate = paste('<b>%{x}</b><br>',
                                'Coverage: %{y}%<br>',
                                '<extra></extra>')) %>%
  layout(
    yaxis = list(
      title = list(text = "Coverage (%)", font = list(size = 16, family = "Arial Black")),
      tickfont = list(size = 14),
      range = c(0, 100),
      dtick = 20
    ),
    xaxis = list(
      title = "",
      tickfont = list(size = 14)
    ),
    showlegend = FALSE
  ) %>%
  animation_opts(frame = 300, transition = 200) %>%
  config(displayModeBar = FALSE)

uk_vaccination_coverage_bar
```

#### (Line view)

```{r}

# Create animated line plot with increased legend font size
uk_vaccination_coverage_line <- plot_ly() %>%
  add_trace(
    data = animated_vac_data,
    x = ~year,
    y = ~coverage,
    color = ~vaccination_type,
    colors = purple_palette,  # Use purple colors
    frame = ~frame,
    type = 'scatter',
    mode = 'lines+markers',
    line = list(width = 3),
    marker = list(size = 8),
    hovertemplate = paste(
      '<b>Year:</b> %{x}<br>',
      '<b>Coverage:</b> %{y}%<br>',
      '<b>%{fullData.name}</b><br>',
      '<extra></extra>'
    )
  ) %>%
  layout(
    xaxis = list(
      title = list(text = "Year", font = list(size = 16, family = "Arial Black")),
      tickfont = list(size = 14),
      range = c(min(years) - 0.5, max(years) + 0.5)
    ),
    yaxis = list(
      title = list(text = "Coverage (%)", font = list(size = 16, family = "Arial Black")),
      tickfont = list(size = 14),
      range = c(0, 100),
      dtick = 20
    ),
    plot_bgcolor = 'white',
    paper_bgcolor = 'white',
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = 'center',
      y = 1.02,
      yanchor = 'bottom',
      bgcolor = 'rgba(255, 255, 255, 0.9)',
      bordercolor = 'rgba(0, 0, 0, 0)',  # Removed border by setting to transparent
      borderwidth = 0,  # Set border width to 0
      itemsizing = "constant",
      itemwidth = 30,  # Increased from 20 to 30 for larger legend box
      font = list(size = 16)  # Increased font size from 14 to 16
    ),
    margin = list(t = 30)
  ) %>%
  animation_opts(
    frame = 300,
    transition = 200,
    redraw = FALSE
  ) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Year: ",
      font = list(size = 16)
    )
  ) %>%
  animation_button(
    x = 0, xanchor = 'left', y = -0.1, yanchor = 'bottom'
  ) %>%
  config(displayModeBar = FALSE)

# Display the plots
uk_vaccination_coverage_line
```

### UK vaccination coverage column 2

```{r}
#| title: "Attitudes of Parents Towards Vaccines in England 2022"


# Load the dataset
uk_parents <- read_xlsx("data/measles_uk/attitudes-of-parents-towards-vaccines-in-england-in-2022.xlsx")

# Define Paired color palette
paired_colors <- brewer.pal(10, "Paired")

# Create static horizontal bar chart
uk_parents_attitudes_bar <- uk_parents %>%
  # Shorten long statements for better display
  mutate(
    statement_short = case_when(
      str_detect(statement, "Measles and mumps continue") ~ "Measles and mumps continue to be a risk",
      str_detect(statement, "Measles can lead to serious complications") ~ "Measles can lead to serious complications",
      str_detect(statement, "Because of vaccinations, smallpox") ~ "Vaccinations have eradicated smallpox",
      str_detect(statement, "Two doses of the MMR vaccine") ~ "Two MMR doses give 99% protection",
      str_detect(statement, "Measles can cause death") ~ "Measles can cause death",
      str_detect(statement, "It doesn't matter if you have missed") ~ "Can catch up on missed MMR doses",
      str_detect(statement, "There is no treatment") ~ "There is no treatment or cure for measles",
      str_detect(statement, "None of the above") ~ "None of the above",
      str_detect(statement, "There were no cases") ~ "No measles/mumps cases in UK this year",
      str_detect(statement, "Measles is harmless") ~ "Measles is harmless",
      TRUE ~ statement
    )
  ) %>%
  plot_ly(
    y = ~reorder(statement_short, percentage_agree), 
    x = ~percentage_agree,
    type = 'bar',
    orientation = 'h',
    color = ~statement_short,
    colors = paired_colors,
    text = ~paste0(percentage_agree, "%"),
    textposition = 'outside',
    hoverinfo = 'none'
  ) %>%
  layout(
    xaxis = list(
      title = list(text = "Percentage Agree (%)", font = list(size = 16, family = "Arial Black")),
      tickfont = list(size = 14),
      range = c(0, 60),
      dtick = 10
    ),
    yaxis = list(
      title = "",
      tickfont = list(size = 12, family = "Arial Black"),
      categoryorder = "array",
      categoryarray = ~reorder(statement_short, percentage_agree),
      tickmode = "array",
      tickvals = ~reorder(statement_short, percentage_agree),
      ticktext = ~statement_short,
      side = "left",
      automargin = TRUE,
      ticklen = 0,
      tickwidth = 0,
      linewidth = 0,
      showgrid = FALSE,
      zeroline = FALSE
    ),
    plot_bgcolor = 'white',
    paper_bgcolor = 'white',
    showlegend = FALSE,
    margin = list(l = 320, r = 50, t = 50, b = 100)  # Increased left margin slightly
  ) %>%
  config(displayModeBar = FALSE)

uk_parents_attitudes_bar
```

## UK ROW 7 - SOURCES {height="300px"}

### 🇬🇧 United Kingdom Data Sources

**Geographic Data**

-   **Source**: UK Regions Topojson Data

-   **URL**: https://onsvisual.github.io/uk-topojson/

-   **Institution**: ONS Visual (Office for National Statistics)

-   **Usage**: Interactive mapping of UK regions for case distribution

**UK Cases by Region and Age**

-   **File**: `cases_region_2012_2024.xlsx`

-   **Source**: [Confirmed cases of measles in England and Wales by region and age 2012-2014](https://www.gov.uk/government/publications/measles-confirmed-cases/confirmed-cases-of-measles-in-england-and-wales-by-region-and-age-2012-to-2014)

-   **Institution**: UK Government / UK Health Security Agency (UKHSA)

-   **Coverage**: Regional breakdown of confirmed measles cases from 2012-2024

-   **Demographics**: Age-stratified data for epidemiological analysis

**UK Vaccination Coverage - Primary Immunisation**

-   **File**: `measles_vac_primary_england_2003_2024.xlsx`

-   **Source**: MMR primary immunisation in England 2024

-   **Institution**: Statista (compiled from NHS England data)

-   **Coverage**: Primary MMR vaccination rates across England (2003-2024)

-   **Purpose**: Tracking first dose vaccination coverage trends

**UK Vaccination Coverage - Complete Immunisation**

-   **File**: `measles_vac_complete_england_2003_2024.xlsx`

-   **Source**: MMR immunization England 2024

-   **Institution**: Statista (compiled from NHS England data)

-   **Coverage**: Complete MMR vaccination rates across England (2003-2024)

-   **Purpose**: Monitoring full vaccination series completion

**UK Parental Attitudes Survey**

-   **File**: `attitudes-of-parents-towards-vaccines-in-england-in-2022.xlsx`

-   **Source**: Attitudes of parents towards vaccines in England 2022

-   **Institution**: Statista (survey data)

-   **Coverage**: Public perception and attitudes toward MMR vaccination

-   **Purpose**: Understanding vaccine hesitancy patterns

# USA

## USA Row 1 - Boxes {height="250px"}

```{r}
#| include: false
#| label: usa_boxes1
#| message: false

# USA DATA ----
usa_cases <- read_xlsx("data/measles_usa/new-cases-of-measles-in-the-us-1985-2025.xlsx") # bar/line chart
usa_age_24 <- read_xlsx("data/measles_usa/cases_by_age_2024.xlsx") # bar chart
usa_age_25 <- read_xlsx("data/measles_usa/cases_by_age_2025.xlsx") # bar chart

# USA States latitude and longitude coordinates
# Source: [Geodatos](https://www.geodatos.net/en/coordinates/united-states), [LatLong.net](https://www.latlong.net/category/states-236-14.html), [Where Am I](https://where-am-i.org/district-of-columbia-latitude.php)
state_coords <- tibble(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
            "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois",
            "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
            "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
            "New Hampshire", "New Jersey", "New Mexico", "New York", "New York City", "North Carolina",
            "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
            "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
            "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  lat = c(32.3182, 66.1605, 34.0489, 34.7999, 36.7783, 39.1130, 41.5999, 39.0000, 38.8950, 27.9944,
          33.2479, 19.7418, 44.0682, 40.0000, 40.2735, 42.0329, 38.5000, 37.8393, 30.3918, 45.3676,
          39.0458, 42.4072, 44.1822, 46.3924, 33.0000, 38.5739, 46.9653, 41.5000, 39.8760, 44.0000,
          39.8339, 34.3071, 43.0000, 40.7143, 35.7822, 47.6506, 40.3675, 36.0846, 44.0000, 41.2033,
          41.7423, 33.8361, 44.5000, 35.8601, 31.0000, 39.4192, 44.0000, 37.9269, 47.7511, 39.0000,
          44.5000, 43.0759),
  lon = c(-86.9023, -153.3691, -111.0937, -92.1999, -119.4179, -105.3589, -72.6999, -75.5000, -77.03667,
          -81.7603, -83.4412, -155.8444, -114.7420, -89.0000, -86.1269, -93.5815, -98.0000, -84.2700,
          -92.3291, -68.9722, -76.6413, -71.3824, -84.5068, -94.6362, -90.0000, -92.6038, -109.5337,
          -100.0000, -117.2241, -71.5000, -74.8718, -106.0181, -75.0000, -74.0060, -80.7935, -100.4370,
          -82.9962, -96.9214, -120.5000, -77.1945, -71.7423, -81.1637, -100.0000, -86.6602, -100.0000,
          -111.9507, -72.6999, -78.0249, -120.7401, -80.5000, -89.5000, -107.2903))

usa_state_24 <- read_xlsx("data/measles_usa/cases_state_2024.xlsx") %>%
  left_join(state_coords, by = "state")
# map + table of top 10

usa_state_25 <- read_xlsx("data/measles_usa/cases_state_2025.xlsx") %>%
  left_join(state_coords, by = "state")
# map + table of top 10

usa_vac_20_25 <- read_xlsx("data/measles_usa/vac_20_25.xlsx") %>%
  mutate(cases = as.integer(cases))  # Convert cases to whole numbers
# stacked bar chart

usa_autism <- read_xlsx("data/measles_usa/us-adults-who-believed-vaccines-cause-autism-2015-2024.xlsx") 
# show this as a pie chart/ stacked bar chart = toggle between

# Calculate key metrics for USA dashboard
total_usa_cases <- sum(usa_cases$cases, na.rm = TRUE)

# Year-over-year change (2025 vs 2024)
cases_2025 <- usa_cases$cases[usa_cases$year == "2025"]
cases_2024 <- usa_cases$cases[usa_cases$year == "2024"]
cases_pct_change <- round(((cases_2025 - cases_2024) / cases_2024) * 100, 1)

# Find peak year
peak_year_data <- usa_cases[which.max(usa_cases$cases), ]

# Find highest state in 2025
highest_state_2025 <- usa_state_25[which.max(usa_state_25$cases), ]

# Most affected age group (combining 2024 and 2025)
combined_age_data <- bind_rows(
  usa_age_24 %>% select(age_group = `Age Group`, cases = Count),
  usa_age_25 %>% select(age_group, cases)
)

age_totals <- combined_age_data %>%
  group_by(age_group) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE)) %>%
  arrange(desc(total_cases))
highest_age_group <- age_totals[1, ]

# Children under 5 percentage (from recent age data)
children_under_5_cases <- combined_age_data %>%
  filter(str_detect(age_group, "Under 5|0-|less than 1|1 to 4")) %>%
  summarise(total = sum(cases, na.rm = TRUE)) %>%
  pull(total)

total_age_cases <- sum(combined_age_data$cases, na.rm = TRUE)
children_under_5_pct <- round((children_under_5_cases / total_age_cases) * 100)

# Average annual cases
avg_annual_cases <- round(total_usa_cases / nrow(usa_cases))

# Unvaccinated percentage in 2025
vac_2025 <- usa_vac_20_25 %>% filter(year == "2025")
total_vac_2025 <- sum(vac_2025$cases, na.rm = TRUE)
unvaccinated_2025 <- vac_2025 %>%
  filter(str_detect(vaccinine_status, "Unvaccinated")) %>%
  summarise(total = sum(cases, na.rm = TRUE)) %>%
  pull(total)
unvaccinated_pct <- round((unvaccinated_2025 / total_vac_2025) * 100)
```

### boxRow 1

```{r}
#| content: valuebox
#| title: "Total Measles Cases (1985-2025)"
#| icon: virus
#| color: danger
list(
  value = paste0(format(total_usa_cases, big.mark = ","), " (+", cases_pct_change, "% in 2025)")
)
```

### boxRow 2

```{r}
#| content: valuebox
#| title: "Highest State Cases (2025)"
#| icon: virus
#| color: danger
list(
  value = paste0(highest_state_2025$state, " - ", format(highest_state_2025$cases, big.mark = ","), " cases")
)
```

### boxRow 3

```{r}
#| content: valuebox
#| title: "Most Affected Age Group"
#| icon: person-fill
#| color: warning
list(
  value = paste0(highest_age_group$age_group, " - ", format(highest_age_group$total_cases, big.mark = ","), " cases")
)
```

### boxRow 4

```{r}
#| content: valuebox
#| title: "Peak Outbreak Year"
#| icon: arrow-up-right-square
#| color: info
list(
  value = paste0(peak_year_data$year, " - ", format(peak_year_data$cases, big.mark = ","), " cases")
)
```

### boxRow 6

```{r}
#| content: valuebox
#| title: "Unvaccinated Cases (2025)"
#| icon: shield-lock
#| color: info
list(
  value = paste0(unvaccinated_pct, "% unvaccinated")
)
```

## USA ROW 2 - STATES {height="1000px"}

### USA cases by Geographic Distribution

```{r}
#| title: Measles Cases by State

# First, create state abbreviation lookup and add to data
state_abbrev <- tibble(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
            "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois",
            "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
            "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
            "New Hampshire", "New Jersey", "New Mexico", "New York", "New York City", "North Carolina",
            "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
            "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
            "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL",
                   "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV",
                   "NH", "NJ", "NM", "NY", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX",
                   "UT", "VT", "VA", "WA", "WV", "WI", "WY")
) %>%
  rename(state_code = abbreviation)

usa_state_24 <- read_xlsx("data/measles_usa/cases_state_2024.xlsx") %>% 
  left_join(state_abbrev, by = "state") %>%
  mutate(year = 2024)

usa_state_25 <- read_xlsx("data/measles_usa/cases_state_2025.xlsx") %>% 
  left_join(state_abbrev, by = "state") %>%
  mutate(year = 2025)

# Combine datasets
combined_data <- bind_rows(usa_state_24, usa_state_25)

# Create color palettes for each year
usa_2024 <- "#0000cd"
usa_2025 <- "#ff4500"

## bar and line charts ----
# Create bar chart with dropdown toggle - Enhanced version
usa_24_25_states_bar_chart <- plot_ly() %>%
  add_bars(
    data = usa_state_24,
    x = ~reorder(state, -cases),
    y = ~cases,
    marker = list(color = usa_2024),
    name = "2024",
    visible = TRUE,
    hovertemplate = paste(
      "<b>%{x}</b><br>",
      "Cases: %{y}<br>",
      "Year: 2024",
      "<extra></extra>"
    )
  ) %>%
  add_bars(
    data = usa_state_25,
    x = ~reorder(state, -cases),
    y = ~cases,
    marker = list(color = usa_2025),
    name = "2025",
    visible = FALSE,
    hovertemplate = paste(
      "<b>%{x}</b><br>",
      "Cases: %{y}<br>",
      "Year: 2025",
      "<extra></extra>"
    )
  ) %>%
  layout(
    xaxis = list(
      title = "State",
      tickangle = -45,
      tickfont = list(size = 14)
    ),
    yaxis = list(
      title = "Number of Cases",
      tickfont = list(size = 14)
    ),
    updatemenus = list(
      list(
        type = "dropdown",
        direction = "down",
        showactive = TRUE,
        x = 0,
        y = 1.1,
        xanchor = "left",
        yanchor = "top",
        bgcolor = "white",
        bordercolor = "gray",
        borderwidth = 1,
        font = list(size = 18),
        buttons = list(
          list(
            label = "2024",
            method = "update",
            args = list(
              list(visible = c(TRUE, FALSE)),
              list()
            )
          ),
          list(
            label = "2025",
            method = "update",
            args = list(
              list(visible = c(FALSE, TRUE)),
              list()
            )
          )
        )
      )
    ),
    margin = list(l = 80, r = 50, t = 120, b = 120),
    showlegend = FALSE
  ) %>%
  config(
    displayModeBar = FALSE
  )

# Display the bar chart
usa_24_25_states_bar_chart
```

### usa maps {.tabset}

```{r}
#| title: USA Measles Cases (2024)

state_abbrev <- tibble(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
            "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois",
            "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
            "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
            "New Hampshire", "New Jersey", "New Mexico", "New York", "New York City", "North Carolina",
            "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
            "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
            "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL",
                   "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV",
                   "NH", "NJ", "NM", "NY", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX",
                   "UT", "VT", "VA", "WA", "WV", "WI", "WY")
) %>%
  rename(state_code = abbreviation)

# Add state codes to your data
usa_state_24 <- read_xlsx("data/measles_usa/cases_state_2024.xlsx") %>% 
  left_join(state_abbrev, by = "state") %>%
  left_join(state_coords, by = "state")

usa_state_25 <- read_xlsx("data/measles_usa/cases_state_2025.xlsx") %>% 
  left_join(state_abbrev, by = "state") %>%
  left_join(state_coords, by = "state")

# USA Maps ----
# Custom Color Gradient for US Measles Cases
us_cases_colors <- list(
  c(0, "#f7fcf5"),        # Pale green (almost white-green)
  c(0.25, "#c7e9c0"),     # Light natural green
  c(0.5, "#74c476"),      # Medium natural green
  c(0.75, "#31a354"),     # Darker natural green
  c(1.0, "#006d2c")       # Deep forest green
)

# Create 2024 Map
us_map_2024 <- plot_geo(usa_state_24, locationmode = "USA-states") %>%
  add_trace(
    locations = ~state_code,  # Keep state codes for correct mapping
    z = ~cases,
    zmin = 0,
    zmax = 100,
    color = ~cases,
    colorscale = us_cases_colors,
    hoverinfo = "text",
    hovertemplate = paste(
      "<b>%{text}</b><br>",
      "Cases: %{z}<br>",
      "Year: %{customdata}",
      "<extra></extra>"
    ),
    text = ~state,  # Ensure state names appear in hover
    customdata = ~year,  # Dynamically show the year
    marker = list(
      line = list(color = "black", width = 0.8)
    ),
    colorbar = list(
      title = list(text = "Measles Cases", side = "bottom"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.1,
      len = 0.8,
      thickness = 15
    )
  ) %>%
  layout(
    geo = list(
      scope = "usa",
      showframe = FALSE,
      showcoastlines = TRUE,
      showland = TRUE,
      landcolor = "white",
      countrycolor = "lightgray",
      projection = list(type = "albers usa"),
      bgcolor = "rgba(0,0,0,0)"
    ),
    margin = list(l = 20, r = 20, t = 20, b = 60)
  ) %>%
  config(displayModeBar = FALSE)

# Create 2025 Map
us_map_2025 <- plot_geo(usa_state_25, locationmode = "USA-states") %>%
  add_trace(
    locations = ~state_code,  
    z = ~cases,
    zmin = 0,
    zmax = 110,
    color = ~cases,
    colorscale = us_cases_colors,
    hoverinfo = "text",
    hovertemplate = paste(
      "<b>%{text}</b><br>",
      "Cases: %{z}<br>",
      "Year: %{customdata}",
      "<extra></extra>"
    ),
    text = ~state,  # Show state names when hovering
    customdata = ~year,
    marker = list(
      line = list(color = "black", width = 1)
    ),
    colorbar = list(
      title = list(text = "Measles Cases", side = "bottom"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.1,
      len = 0.8,
      thickness = 15,
      tick0 = 0,
      dtick = 20,
      tickvals = c(0, 20, 40, 60, 80, 100),
      ticktext = c("0", "20", "40", "60", "80", "100+")
    )
  ) %>%
  layout(
    geo = list(
      scope = "usa",
      showframe = FALSE,
      showcoastlines = TRUE,
      showland = TRUE,
      landcolor = "white",
      countrycolor = "lightgray",
      projection = list(type = "albers usa"),
      bgcolor = "rgba(0,0,0,0)"
    ),
    margin = list(l = 20, r = 20, t = 20, b = 60)
  ) %>%
  config(displayModeBar = FALSE)

us_map_2024
```

```{r}
#| title: (2025)
us_map_2025
```

## USA ROW 3 - Vac boxes {height="250px"}

```{r}
#| include: false
#| label: usa_boxes2
#| message: false
usa_vac_20_25 <- read_xlsx("data/measles_usa/vac_20_25.xlsx") %>%
  mutate(cases = as.integer(cases))  # Convert cases to whole numbers
usa_autism <- read_xlsx("data/measles_usa/us-adults-who-believed-vaccines-cause-autism-2015-2024.xlsx") 
# Vaccination effectiveness metrics
total_unvaccinated <- sum(usa_vac_20_25$cases[usa_vac_20_25$vaccinine_status == "Unvaccinated or Unknown"], na.rm = TRUE)
total_vac_cases <- sum(usa_vac_20_25$cases, na.rm = TRUE)
unvaccinated_pct <- (total_unvaccinated / total_vac_cases) * 100
# Two-dose vaccination cases
total_two_dose <- sum(usa_vac_20_25$cases[usa_vac_20_25$vaccinine_status == "Two MMR doses"], na.rm = TRUE)
two_dose_pct <- (total_two_dose / total_vac_cases) * 100
vaccine_effectiveness <- 100 - two_dose_pct
# Autism belief trends
autism_2015 <- usa_autism %>% filter(year == 2015)
autism_2024 <- usa_autism %>% filter(year == 2024)
# Calculate belief percentages
yes_cause_2015 <- autism_2015$distribution[autism_2015$autism == "Yes, a cause"] / sum(autism_2015$distribution) * 100
yes_cause_2024 <- autism_2024$distribution[autism_2024$autism == "Yes, a cause"] / sum(autism_2024$distribution) * 100
belief_change <- yes_cause_2024 - yes_cause_2015
# Unsure percentage in 2024
unsure_2024 <- autism_2024$distribution[autism_2024$autism == "Unsure"] / sum(autism_2024$distribution) * 100
```

### box vac1

```{r}
#| content: valuebox
#| title: "Unvaccinated Case Rate (2020-2025)"
#| icon: shield-lock
#| color: warning
#| fig-height: 200
list(
  value = paste0(round(unvaccinated_pct, 0), "% (", 
                total_unvaccinated, "/", total_vac_cases, ")")
)
```

### box vac2

```{r}
#| content: valuebox
#| title: "Vaccination Effectiveness"
#| icon: shield-check
#| color: success
#| fig-height: 200
list(
  value = paste0(round(vaccine_effectiveness, 0), "% reduction in cases")
)
```

## USA Row 4 - Vaccination donuts {height="1000px"}

### USA Measles Cases by Vaccination Status {.tabset}

#### USA Measles Cases by Vaccination Status 2020

```{r}

# Calculate percentages
usa_vac_percentage <- usa_vac_20_25 %>%
  group_by(year) %>%
  mutate(total_cases = sum(cases),
         percentage = round((cases / total_cases) * 100, 0)) %>%
  ungroup()

# Define colors for consistency
colors <- c("Unvaccinated or Unknown" = "#738",
            "One MMR dose" = "#0A7029", 
            "Two MMR doses" = "#52688F")

# Modified Donut Chart Function
create_donut_chart <- function(year_data, year_title) {
  # Filter out zero values for cleaner visualization
  year_data_filtered <- year_data %>% filter(cases > 0)
  
  if (nrow(year_data_filtered) == 0) {
    return(NULL)
  }
  
  plot_ly(
    data = year_data_filtered,
    labels = ~vaccinine_status,  # Use actual column name from dataset
    values = ~cases,
    type = 'pie',
    hole = 0.5,
    textinfo = 'value+percent',  # Show both cases and percentages
    textposition = 'inside',  # Place text inside the donut segments
    textfont = list(size = 16, color = "white"),  # White text for better contrast
    texttemplate = '%{value}<br>(%{percent})',  # Format: cases on top, percentage below in parentheses
    hovertemplate = paste(
      '<b>%{label}</b><br>',
      'Cases: %{value}<br>',
      'Percentage: %{percent}<br>',
      '<extra></extra>'
    ),
    marker = list(
      colors = colors[year_data_filtered$vaccinine_status],
      line = list(color = 'white', width = 1)
    ),
    showlegend = TRUE,
    domain = list(x = c(0, 1), y = c(0, 0.85))
  ) %>%
    layout(
      font = list(family = "Arial, sans-serif", size = 12),
      margin = list(t = 100, b = 40, l = 40, r = 40),
      legend = list(
        orientation = "h",
        x = 0.5,
        y = 1,
        xanchor = "center",
        yanchor = "bottom",
        font = list(size = 16) # Change this to 16
      ),
      annotations = list(
        list(
          text = paste0("<b>", year_title, "</b>"),
          x = 0.5, y = 0.5,
          font = list(size = 20, color = "black"),
          showarrow = FALSE
        ),
        list(
          text = paste0("Total Cases<br><b>", sum(year_data$cases), "</b>"),
          x = 0.5, y = 0.4,
          font = list(size = 20, color = "black"),
          showarrow = FALSE))
    ) %>%
    config(displayModeBar = FALSE)
}

# Alternative function with better text positioning for small segments
create_donut_chart_smart <- function(year_data, year_title) {
  year_data_filtered <- year_data %>% filter(cases > 0)
  
  if (nrow(year_data_filtered) == 0) {
    return(NULL)
  }
  
  # Determine text position based on percentage size
  year_data_filtered <- year_data_filtered %>%
    mutate(
      text_position = ifelse(percentage >= 15, "inside", "outside"),
      text_info = ifelse(percentage >= 15, "value+percent", "label+value+percent")
    )
  
  plot_ly() %>%
    add_pie(
      data = year_data_filtered,
      labels = ~vaccinine_status,
      values = ~cases,
      hole = 0.4,
      textinfo = 'value+percent',
      textposition = ~text_position,
      texttemplate = '%{value}<br>(%{percent})',  # Format: cases on top, percentage in parentheses
      textfont = list(size = 14),
      hovertemplate = paste(
        '<b>%{label}</b><br>',
        'Cases: %{value}<br>',
        'Percentage: %{percent}<br>',
        '<extra></extra>'
      ),
      marker = list(
        colors = colors[year_data_filtered$vaccinine_status],
        line = list(color = 'white', width = 1)
      ),
      showlegend = TRUE,
      domain = list(x = c(0, 1), y = c(0, 0.85))
    ) %>%
    layout(
      font = list(family = "Arial, sans-serif", size = 12),
      margin = list(t = 100, b = 40, l = 40, r = 40),
      legend = list(
        orientation = "h",
        x = 0.5,
        y = 1,
        xanchor = "center",
        yanchor = "bottom",
        font = list(size = 16) # Change this to 16
      ),
      annotations = list(
        list(
          text = paste0("<b>", year_title, "</b>"),
          x = 0.5, y = 0.5,
          font = list(size = 20, color = "black"),
          showarrow = FALSE
        ),
        list(
          text = paste0("Total Cases<br><b>", sum(year_data$cases), "</b>"),
          x = 0.5, y = 0.4,
          font = list(size = 20, color = "black"),
          showarrow = FALSE
        )
      )
    ) %>%
    config(displayModeBar = FALSE)
}

# Create individual donut charts for each year
years <- unique(usa_vac_percentage$year)
donut_plots <- list()

for (year in years) {
  year_data <- usa_vac_percentage %>% filter(year == !!year)
  donut_plots[[as.character(year)]] <- create_donut_chart_smart(year_data, year)
}

donut_plots[["2020"]]
```

#### 2021

```{r}
donut_plots[["2021"]]
```

#### 2022

```{r}
donut_plots[["2022"]]
```

#### 2023

```{r}
donut_plots[["2023"]]
```

#### 2024

```{r}
donut_plots[["2024"]]
```

#### 2025

```{r}
donut_plots[["2025"]]
```

### USA Beliefs About Vaccines Causing Autism {.tabset}

#### USA Beliefs About Vaccines Causing Autism 2015

```{r}
# Load required libraries
library(plotly)
library(dplyr)

# Define colors for beliefs
autism_colors <- c("Yes, a cause" = "#d62728",  # Red
            "No, not a cause" = "#2ca02c",  # Green
            "Unsure" = "#ff7f0e")  # Orange

# Function to create donut chart
create_autism_donut_chart <- function(year_data, year_title) {
  # Filter out zero values for cleaner visualization
  year_data_filtered <- year_data %>% filter(distribution > 0)

  if (nrow(year_data_filtered) == 0) {
    return(NULL)
  }

  plot_ly(
    data = year_data_filtered,
    labels = ~autism,
    values = ~distribution,
    type = 'pie',
    hole = 0.5,
    textinfo = 'value+percent',
    textposition = 'inside',
    textfont = list(size = 16, color = "white"),
    texttemplate = '%{value}<br>(%{percent})',
    hovertemplate = paste(
      '<b>%{label}</b><br>',
      'Responses: %{value}<br>',
      'Percentage: %{percent}<br>',
      '<extra></extra>'
    ),
    marker = list(
      colors = autism_colors[year_data_filtered$autism],
      line = list(color = 'white', width = 1)
    ),
    showlegend = TRUE,
    domain = list(x = c(0, 1), y = c(0, 0.85))
  ) %>%
    layout(
      font = list(family = "Arial, sans-serif", size = 12),
      margin = list(t = 100, b = 40, l = 40, r = 40),
      legend = list(
        orientation = "h",
        x = 0.5,
        y = 1,
        xanchor = "center",
        font = list(size = 16)
      ),
      annotations = list(
        list(
          text = paste0("<b>", year_title, "</b>"),
          x = 0.5, y = 0.5,
          font = list(size = 20, color = "black"),
          showarrow = FALSE
        ),
        list(
          text = paste0("Total Responses<br><b>", sum(year_data$distribution), "</b>"),
          x = 0.5, y = 0.4,
          font = list(size = 20, color = "black"),
          showarrow = FALSE
        )
      )
    ) %>%
    config(displayModeBar = FALSE)
}

# Generate donut charts for each year
autism_years <- unique(usa_autism$year)
donut_plots_autism <- list()

for (year in autism_years) {
  year_data <- usa_autism %>% filter(year == !!year)
  donut_plots_autism[[as.character(year)]] <- create_autism_donut_chart(year_data, year)
}

donut_plots_autism[["2015"]]
```

#### 2019

```{r}
#| title: (2019)
donut_plots_autism[["2019"]]
```

#### 2024

```{r}
#| title: (2024)
donut_plots_autism[["2024"]]
```

## USA Row 5 - Vaccination Summary {height="800px"}

### usa final row col 1

```{r}
#| title: MMR Vaccination Status and Measles Cases (2020-2025)

timeline_data <- usa_vac_20_25 %>%
  mutate(year_num = as.numeric(year)) %>%
  arrange(year_num, vaccinine_status)

timeline_bar <- timeline_data %>%
  plot_ly(
    x = ~vaccinine_status,
    y = ~cases,
    color = ~vaccinine_status,
    colors = colors,
    type = 'bar',
    frame = ~year,
    text = ~cases,
    textposition = 'outside',
    hovertemplate = paste(
      '<b>%{x}</b><br>',
      'Year: %{frame}<br>',
      'Cases: %{y}<br>',
      '<extra></extra>'
    )
  ) %>%
  layout(
    xaxis = list(
      title = "<b>Vaccination Status</b>",
      font = list(size = 16),
      tickfont = list(size = 14)  # Increase axis value text size
    ),
    yaxis = list(
      title = "<b>Number of Cases</b>",
      range = c(0, max(timeline_data$cases) + 15),
      font = list(size = 16),
      tickfont = list(size = 14)  # Increase axis value text size
    ),
    showlegend = FALSE
  ) %>%
  animation_opts(
    frame = 400,
    transition = 300,
    easing = "cubic-in-out",
    redraw = FALSE
  ) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Year: ",
      font = list(size = 16, color = "#333")
    ),
    len = 0.8,
    x = 0.1
  ) %>%
  animation_button(
    x = 0.1, y = 0,
    bgcolor = "white",
    bordercolor = "black",
    font = list(color = "black")
  ) %>%
  config(
    displayModeBar = FALSE,  # Completely hides the mode bar
    staticPlot = TRUE        # Disables all interactive features
  )

timeline_bar
```

### usa final row col 2 {width="40%"}

```{r}
#| title: MMR Vaccination Status and Measles Cases

summary_year <- usa_vac_20_25%>%
  rename(Year = year, 
         `Vaccination Status` = vaccinine_status, 
         Cases = cases)

us_cases_colors2 <- list("#f7fcf5", # Pale green (almost white-green) 
                        "#c7e9c0", # Light natural green 
                        "#74c476" # Medium natural green 
                        )

# Define breakpoints for the color intervals
breaks <- quantile(summary_year$Cases, probs = seq(0, 1, length.out = 3), na.rm = TRUE)

# Create the interactive table with updated column names and color styling
final_table <- datatable(summary_year, 
                         rownames = FALSE,
                         class = "compact stripe hover order-column",
                         options = list(
                           pageLength = 18,
                           lengthMenu = list(c(9, 12, 16, 18), c("9", "12", "16", "All")),
                           autoWidth = FALSE,
                           columnDefs = list(
                             list(targets = "_all", className = "dt-left")
                           )
                         ),
                         filter = "top"
) %>%
  formatStyle(
    names(summary_year),
    fontFamily = "Arial",
    fontSize = "14px",
    textAlign = "left"
  ) %>%
  formatStyle(
    "Cases",
    backgroundColor = styleInterval(breaks[-1], us_cases_colors2),
    fontWeight = "bold"
  )

# Display the updated table
final_table
```

## USA ROW 6 - SOURCES {height="300px"}

### 🇺🇸 United States Data Sources

**USA State Codes**

-   **Source**: US State Abbreviations

-   **URL**: https://www.50states.com/abbreviations.htm

-   **Usage**: Standardised state coding for mapping and data processing

**USA Historical Cases Trend**

-   **File**: `new-cases-of-measles-in-the-us-1985-2025.xlsx`

-   **Source**: [New cases of measles in the US since 1950](https://www.statista.com/statistics/186678/new-cases-of-measles-in-the-us-since-1950/)

-   **Institution**: Statista (compiled from CDC data)

-   **Coverage**: Long-term trend analysis from 1985-2025

-   **Purpose**: Historical context and outbreak pattern analysis

**USA Cases by Age Group (2024)**

-   **File**: `cases_by_age_2024.xlsx`

-   **Source**: [Number of measles cases by age U.S. 2020-2024](https://www.statista.com/statistics/1469710/number-measles-cases-in-the-us-by-age/)

-   **Institution**: Statista (CDC surveillance data)

-   **Coverage**: Age-stratified case distribution for 2024

-   **Purpose**: Identifying vulnerable age groups

**USA Cases by Age Group (2025)**

-   **File**: `cases_by_age_2025.xlsx`

-   **Source**: [Number of measles cases by age U.S. 2024-2025](https://www.statista.com/statistics/1560807/number-measles-cases-by-age/)

-   **Institution**: Statista (CDC surveillance data)

-   **Coverage**: Current age-stratified case distribution

-   **Purpose**: Real-time epidemiological monitoring

**USA Cases by State (2024)**/

-   **File**: `cases_state_2024.xlsx`

-   **Source**: Number of measles by state U.S. 2024

-   **Institution**: Statista (CDC surveillance data)

-   **Coverage**: State-level case distribution for 2024

-   **Purpose**: Geographic outbreak mapping

**USA Cases by State (2025)**

-   **File**: `cases_state_2025.xlsx`

-   **Source**: Number of measles by state U.S. 2025

-   **Institution**: Statista (CDC surveillance data)

-   **Coverage**: Current state-level case distribution

-   **Purpose**: Real-time outbreak monitoring

**USA Vaccination Status Analysis**

-   **Files**:

    -   `vac_2024.xlsx` - Number of measles cases by vaccination status U.S. 2020-2024

    -   `vac_2025.xlsx` - Distribution of measles cases by vaccination status U.S. 2024-2025

    -   `vac_20_25.xlsx` - Combined vaccination status data 2020-2025

-   **Institution**: Statista (CDC surveillance data)

-   **Coverage**: Vaccination status of confirmed measles cases

-   **Purpose**: Vaccine effectiveness monitoring and breakthrough case analysis

**US Public Perception Survey**

-   **File**: `us-adults-who-believed-vaccines-cause-autism-2015-2024.xlsx`

-   **Source**: Opinions on whether vaccines cause autism U.S. 2024

-   **Institution**: Statista (survey research)

-   **Coverage**: Public beliefs about MMR-autism connection (2015-2024)

-   **Purpose**: Understanding vaccine hesitancy drivers

# About Measles

### 📌 Measles Overview

**Measles**

**Description**: Measles is one of the most contagious viral diseases, caused by the Morbillivirus.

**Transmission Rate**: One infected person can infect 12–18 others in a susceptible population.

**Contagious Period**: 4 days before to 4 days after rash onset.

**Clinical Recognition**: \* **Days 1–3**: Fever, cough, runny nose, conjunctivitis \* **Days 2–3**: Koplik’s spots (inside mouth) \* **Days 3–5**: Red, blotchy rash starts at hairline, spreads downward

**Public Health Note**: Early detection is critical, as patients are contagious before symptoms fully appear.

**Sources**: \* [CDC Measles Clinical Features](https://www.cdc.gov/measles/about/index.html) \* [WHO Measles Fact Sheet](https://www.who.int/news-room/fact-sheets/detail/measles)

### 📌 Understanding the Disease

**What is Measles**

**Cause**: Measles virus, a highly contagious respiratory virus.

**Transmission Potential**: R₀ = 12–18.

**Historical Impact**: Pre-vaccine, caused \~2.6 million deaths annually.

**Outbreak Relevance**: Continues to emerge in areas with low vaccination coverage.

**Sources**: \* [CDC Clinical Features](https://www.cdc.gov/measles/about/index.html) \* [WHO Fact Sheet](https://www.who.int/news-room/fact-sheets/detail/measles)

### 📌 From Ancient Scourge to Vaccine-Preventable Disease

**Evolutionary Origins & Historical Spread**

**Origin**: Likely evolved from rinderpest \~11th–12th centuries.

**Population Dependency**: Requires high population density to sustain transmission.

**Historical Spread**: Moved with trade and colonial expansion.

**Modern Relevance**: Reflects today's outbreaks in unvaccinated populations.

**Dashboard Insight**: GBD 2021 data shows enduring outbreak patterns.

**Sources**:

-   [Measles Evolution Study](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3905323/)
-   [Historical Epidemiology](https://journals.asm.org/doi/10.1128/JVI.00417-06)

### 📌 Who Is at Risk?

**High-Risk Populations**

-   **Unvaccinated individuals**
-   **Infants \<12 months**
-   **Pregnant women (no immunity)**
-   **Immunocompromised individuals**
-   **International travelers**
-   **Healthcare workers without verified immunity**
-   **Communities with vaccine exemptions**

**Critical Surveillance Points**

-   Airports, emergency departments, schools, healthcare settings

**Dashboard Insights**

-   Breakthrough infections: \~3% of vaccinated
-   Unvaccinated risk: significantly higher during outbreaks

### 📌 Transmission Mechanisms

**Transmission Facts**

-   **Airborne Survival**: Up to 2 hours in air/surfaces
-   **High Infectivity**: 90% of susceptible contacts become infected

**Infectious Timeline**: \* **Incubation**: 7–21 days (avg 10–14) \* **Peak Infectiousness**: Prodromal stage

**R₀ Estimate**: 12–18 **Immunity Threshold**: 95% needed to prevent spread

**Sources**: \* [CDC Transmission Guidelines](https://www.cdc.gov/measles/transmission/index.html) \* [WHO Epidemiology Review](https://www.who.int/immunization/monitoring_surveillance/burden/vpd/WHO_SurveillanceVaccinePreventable_11_Measles_R2.pdf)

### 📌 Vaccine Characteristics & Implementation

**Vaccine Profile**

-   **Type**: Live attenuated (MMR)
-   **Components**: Measles, Mumps, Rubella
-   **Storage**: Frozen (-15°C or colder)
-   **Administration**: Subcutaneous injection

**Schedule & Considerations**

-   **Dose 1**: 12–15 months
-   **Dose 2**: 4–6 years
-   **Catch-Up**: For adults born 1957+, healthcare workers, travelers
-   **Contraindicated In**: Pregnant or immunocompromised individuals
-   **Dashboard Insights**: WHO/UNICEF data shows regions below 90% MCV1 coverage have frequent outbreaks.

**Source**: [Our World in Data](https://ourworldindata.org/measles)

### 📌 Vaccine Effectiveness & Coverage Targets

**Effectiveness Data**

-   **Single Dose**: 93% effective
-   **Two Doses**: 97% effective
-   **Duration**: Lifelong immunity in most cases

**Coverage Goals**

-   **Target**: 95% two-dose coverage
-   **Outbreak Risk**: Transmission possible below 90–92%
-   **Herd Immunity**: Needs consistently high coverage across all areas

**Contraindications**: Pregnancy, severe immunodeficiency, recent blood products, severe illness, allergy to components

**Real-World Evidence**: UK MMR coverage (2003–2024) shows correlation between coverage dips and outbreaks

-   **Sources**:
    -   [CDC MMR Info](https://www.cdc.gov/vaccines/vpd/mmr/index.html)
    -   [WHO MMR Position Paper](https://www.who.int/publications/i/item/WER9017)

### 📌 Safety Profile & Adverse Events

**Common Side Effects**

-   **Local**: Redness, soreness at injection site (10–15%)
-   **Systemic**: Fever (5–15%), mild rash (5%)
-   **Timing**: 7–12 days post-vaccination

**Moderate Events**

-   Febrile seizures (1 in 3,000–4,000 doses)
-   Thrombocytopenia (1 in 30,000–40,000)
-   Joint pain (adults, especially women)

**Severe Events (Very Rare)**

-   Anaphylaxis: \<1 in 1 million
-   Encephalitis: \<1 in 1 million

**Risk-Benefit Evidence**: GBD 2021 shows disease complications far outweigh vaccine risks, especially in infants and adults.

# Data Sources

### 🌍 Global Measles Data

**Global Measles Surveillance**

**Description**: Provides insights into worldwide measles cases and mortality.

**Key Datasets**:

`global_cases_deaths.csv` – Global case/death data

`global_cases_sex_age.csv` – Age-stratified deaths

`measles_world` – Reported cases

`measles_death` – Historical deaths

**Sources**:

-   [GBD Study 2021](https://vizhub.healthdata.org/gbd-results/)

-   [Our World in Data Measles Tracking](https://ourworldindata.org/grapher/reported-cases-of-measles)

### 💉 Vaccination Coverage Data

**Global Immunization Tracking**

**Purpose**: Tracks country-level immunization rates.

**Coverage Metrics**:

-   MCV1 (1st dose): `1dose_share-of-children-vaccinated-against-measles.csv`

-   MCV2 (2nd dose): `2_doses_share-of-children-vaccinated-with-mcv.csv`

**Validation**: WHO & UNICEF estimates

**Sources**:

-   [MCV1 Coverage](https://ourworldindata.org/grapher/share-of-children-vaccinated-against-measles)

-   [MCV2 Coverage](https://ourworldindata.org/grapher/share-of-children-vaccinated-with-mcv2)

### 🇺🇸 United States Data

**US Measles Surveillance**

**Institution**: CDC (via NNDSS)

**Key Datasets**:

-   `new-cases-of-measles-in-the-us-1985-2025.xlsx` – Historical trends

-   `cases_by_age_2024.xlsx`, `cases_by_age_2025.xlsx` – Age distribution

-   `cases_state_2024.xlsx`, `cases_state_2025.xlsx` – State-level data

-   [US State Abbreviations](https://www.50states.com/abbreviations.htm)

-   `vac_2024.xlsx`, `vac_2025.xlsx`, `vac_20_25.xlsx` – Vaccination status

-   `us-adults-who-believed-vaccines-cause-autism-2015-2024.xlsx` – Public beliefs

**Source**: [CDC NNDSS](https://www.cdc.gov/nndss/)

### 🇬🇧 United Kingdom Data

**UKHSA Surveillance**

**Institution**: UK Health Security Agency (UKHSA)

**Key Datasets**:

-   `cases_region_2012_2024.xlsx` – Regional surveillance

-   `measles-cases-in-england-and-wales-2023-and-2024-by-age.xlsx` – Age-specific data

-   `measles_vac_primary_england_2003_2024.xlsx` – Primary vaccine coverage

-   `measles_vac_complete_england_2003_2024.xlsx` – Full coverage data

-   `attitudes-of-parents-towards-vaccines-in-england-in-2022.xlsx` – Public sentiment

-   [ONS TopoJSON Maps](https://onsvisual.github.io/uk-topojson/)

**Source**: [UKHSA](https://www.gov.uk/government/organisations/uk-health-security-agency)

### ⚠️ Data Quality & Limitations

**Important Considerations**

**Underreporting**:

-   Mild/unreported cases
-   Incomplete lab confirmations
-   Delay in reporting

**Coverage Limitations**:

-   Survey vs. admin data inconsistencies
-   Population denominator issues

**Temporal Factors**:

-   Reporting delays
-   Seasonal variations
-   COVID-era disruptions

**Known Gaps**:

-   Infrequent updates
-   Diagnostic variation
-   Systemic weaknesses

### 🧪 Quality Control & Methodology

**Data Integration and Validation**

**Sources**:

-   Our World in Data
-   GBD 2021 (IHME)
-   WHO/UNICEF
-   National reporting systems

**Cross-Validation**:

-   Time-series alignment
-   Multi-source comparison
-   Demographic harmonization

**Quality Assurance**:

-   Standard definitions
-   Peer-reviewed methods
-   Routine updates

**Methodology & Source Reliability**

**Ranking of Reliability**:

1.  National surveillance (CDC, UKHSA)
2.  WHO/UNICEF estimates
3.  GBD modeling (IHME)
4.  Statista/surveys

**Key Indicators**:

-   Timeliness
-   Accuracy
-   Completeness
-   Correction mechanisms

### 📚 Master Dataset Catalogue

**Complete Dataset List**

**Global Data**:

-   `global_cases_deaths.csv`
-   `global_cases_sex_age.csv`
-   `measles_world`, `measles_death`, `measles_death_age`
-   `1dose_share-of-children-vaccinated-against-measles.csv`
-   `2_doses_share-of-children-vaccinated-with-mcv.csv`

**USA Files**:

-   `new-cases-of-measles-in-the-us-1985-2025.xlsx`
-   `cases_by_age_2024.xlsx`, `cases_by_age_2025.xlsx`
-   `cases_state_2024.xlsx`, `cases_state_2025.xlsx`
-   `vac_2024.xlsx`, `vac_2025.xlsx`, `vac_20_25.xlsx`
-   `us-adults-who-believed-vaccines-cause-autism-2015-2024.xlsx`

**UK Files**:

-   `cases_region_2012_2024.xlsx`
-   `measles_vac_primary_england_2003_2024.xlsx`
-   `measles_vac_complete_england_2003_2024.xlsx`
-   `attitudes-of-parents-towards-vaccines-in-england-in-2022.xlsx`

**Geographic Files**:

-   [ONS TopoJSON](https://onsvisual.github.io/uk-topojson/)
-   [US State Codes](https://www.50states.com/abbreviations.htm)

### 📝 Citation Policy

**Dashboard Citation**

``` text
Measles Watch Dashboard (2024). Comprehensive measles outbreak monitoring and vaccination data.  
Available at: https://github.com/KFarrow11/datavis_assignment2
```
